---
title: "Code for website"
output: html_notebook
---

Create a project for a *package*.

```{r}
library(usethis)
library(pkgdown)
```


Build a website for a package.
```{r}
usethis::use_pkgdown() # creates _plgdown.yml file
pkgdown::build_site() # creates docs folder with files for the website

pkgdown::preview_site() # preview the site
```

Add this repository to Github (I used desktop app). Make sure it's public.
Have a look at your GitHub if it's there already. For me it shows that there's a folder only (no contents yet). We need to push the files to Github.

```{r}
# now run the code below to link your website to GitHub pages
use_pkgdown_github_pages()
usethis::use_github_action("pkgdown") 
# pkgdown::preview_site()
```

Now in the GitHub repository's "About" section, you should see your website's link. Your _pkgdown.yml should have a URL of your GitHub Pages.


It might show 404 error. Wait for it. Also, you can change the source of your GitHub Pages if you go to the repo's settings -> Pages -> source
--> Mine is: gh-pages /(root)

Create a README file. Make sure it's knitted as md file (in the Rmd add: output: md_document) or it will not be recognised. Push changes


```{r}
# Might be needed to push the files to GitHub
# usethis::use_git() # agree to push
# usethis::use_github()
```


.Rbuildignore

^.*\.Rproj$
^\.Rproj\.user$
^_pkgdown\.yml$
^docs$
^pkgdown$
^\.github$
^README\.Rmd$


web/ .gitignore   in master

.Rproj.user
.Rhistory
.RData
.Ruserdata
.Rdata
.httr-oauth
.DS_Store
docs



web/.github/.gitignore 
*.html



 
 



