---
output: github_document
bibliography: references.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE
)
library(sf)
library(tmap)
```

```{r, eval=FALSE}
# Generate citations (requires Zotero)
library(rbbt)
bbt_write_bib(path = "references.bib", keys = bbt_detect_citations("README.Rmd"), overwrite = TRUE)
```




# Open access data for transport research: tools, modelling and simulation

<!-- badges: start -->

This repo contains code and example data to explore the utility of open data for transport planning and, specifically, open data on transport infrastructure.
It was created to support a LIDA internship, the objectives of which are to:

    • develop new methods for bulk downloading, querying and analysing OpenStreetMap data on transport infrastructure
    • assess the quality of OSM data with reference to ‘ground truth’ datasets including data from satellite imagery and Ordnance Survey data
    • publish reproducible methods and documentation on using OSM data for transport planning with reference to the strengths and potential pitfalls of the data
    • develop a typology of transport infrastructure data and data schemas for each infrastructure type and an actionable definition of ‘active travel infrastructure’
    • develop ‘OSM transport infrastructure data packs’ for every transport authority in Great Britain, with layers reflecting a typology of transport infrastructure data developed in the project
    • develop and publish guidance on using OSM data for transport planning

# References

