---
output: github_document
bibliography: references.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE
)
library(sf)
library(tmap)
```

```{r, eval=FALSE}
# Generate citations (requires Zotero)
library(rbbt)
bbt_write_bib(path = "references.bib", keys = bbt_detect_citations("README.Rmd"), overwrite = TRUE)
```




# Open access data for transport research: tools, modelling and simulation

<!-- badges: start -->
[![.github/workflows/render-rmarkdown.yaml](https://github.com/udsleeds/openinfra/actions/workflows/render-rmarkdown.yaml/badge.svg)](https://github.com/udsleeds/openinfra/actions/workflows/render-rmarkdown.yaml)
<!-- badges: end -->

Origin-destination (OD) datasets are ubiquitous for representing travel behavior in transport planning and modelling.
Despite emerging large data sources, OD data continues to play an established --- if not central --- role in transport research in the 21^st^ century, with recent applications ranging from analysis of the evolution of urban activity and shared mobility services over time [e.g. @shi_exploring_2019; @li_effects_2019] to inference of congestion and mode split [@bachir_inferring_2019; @gao_method_2021]. 

# References

