---
title: "Transport Infrastructure Data Packs"
author: "James Hulse"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Transport Infrastructure Data Packs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Packs

This vignette covers the recategorisation of OpenStreetMap (OSM) infrastructure data into the openinfra transport infrastructure data packs.
<br><br>
The transport infrastructure data packs contain OSM infrastructure data for a given Local Authority District (LAD), as defined by the following polygons for [UK LADs](https://github.com/udsleeds/openinfra/blob/main/data-small/lads_joined_2021.geojson). 
<br><br>
Whilst these LADs are currently being used to define the spatial area covered by each infrastructure data pack, any new set of updated boundaries (as a geojson file) can also be used to create the transport infrastructure data packs. 
<br><br>

The LADs can be seen below:<br>

```{r, echo=FALSE, include=TRUE, out.width="650px"}
knitr::include_url("https://udsleeds.github.io/openinfraresults/tmap/assets/LADs_of_the_UK_new.html", height = "650px")
```

## Current Data Pack Example Visualisations

This sections visualises the data pack outputs generated by the openinfra functions above. 

The current example data pack for "Leeds" - contains outputs: <br>
-- Cleaned road descriptions <br>
-- Active cycle network <br>
-- Active walking network <br>
-- Presence of lighting network <br>
-- Cleaned maximum speed values <br>
-- Inclusive Mobility (unfortunately not mapped here due to map filesize restrictions) 
<br><br>


### Default OSM Highways

Please note that a 7.5Km circular buffer (centred at Leeds City Centre) has been applied to this map so that the map can be hosted on this website. (max filesize 25Mb)<br>
As can be seen, the default network has a variety of `highway` values and is rather confusing. 
```{r out.width="100%", echo=FALSE, include=TRUE}
knitr::include_url("https://udsleeds.github.io/openinfraresults/tmap/assets/deault_OSM_highways.html", height = "650px")
```
<br><br>

### Recategorised Road Descriptions
The recategorised road descriptions are more interpenetrate and meaningful than the default OSM highway values.

```{r visualisation_frame, out.width="100%", echo=FALSE, include=TRUE}
knitr::include_url("https://udsleeds.github.io/openinfraresults/tmap/assets/road_desc_map.html", height = "650px")
```
As can be seen - the OSM highways re-categorised to road descriptions are much easier to visualise and insightful than all OSM highway values.
<br><br>

### Recategorised Cyclable Ways
```{r out.width="100%", echo=FALSE, include=TRUE}
knitr::include_url("https://udsleeds.github.io/openinfraresults/tmap/assets/active_cycle_map.html", height = "650px")
```

<br><br>

### Recategorised Walkable Ways
```{r out.width="100%", echo=FALSE, include=TRUE}
knitr::include_url("https://udsleeds.github.io/openinfraresults/tmap/assets/active_walk_map.html", height = "650px")
```
<br><br>

### Recategorised Presence of Street Lighting
```{r out.width="100%", echo=FALSE, include=TRUE}
knitr::include_url("https://udsleeds.github.io/openinfraresults/tmap/assets/is_lit_map.html", height = "650px")
```
Note that a value of `maybe` implies the lack of data to distinguish between `"yes"` or `"no"`.

<br><br>

### Cleaned Maximum Speeds (UK)
```{r vis_frame, out.width="100%", echo=FALSE, include=TRUE}
knitr::include_url("https://udsleeds.github.io/openinfraresults/tmap/assets/clean_maxspeed_map.html", height = "650px")
```

### Presence of Cycle Crossings
```{r out.width="100%", echo=FALSE, include=TRUE}
knitr::include_url("https://udsleeds.github.io/openinfraresults/tmap/assets/cycle_crossings_map.html", height = "650px")
```

### Presence of Cycle Parking
```{r out.width="100%", echo=FALSE, include=TRUE}
knitr::include_url("https://udsleeds.github.io/openinfraresults/tmap/assets/cycle_parking_map.html", height = "650px")
```

### Recategorised Road Names
```{r out.width="100%", echo=FALSE, include=TRUE}
knitr::include_url("https://udsleeds.github.io/openinfraresults/tmap/assets/road_namesmap.html", height = "650px")
```


## Setup
The following libraries are required for the production of transport infrastructure data packs.
```{r libraries, echo=TRUE, message=FALSE, eval=FALSE}
library(sf)
library(dplyr)
library(tmap)
library(openinfra)
```
<br><br>

### Data Packs: Value Added Data - The Functions

This section covers the current Openinfra (`oi_`) functions utilised to recategorise and add value to OSM infrastructure network data. <br><br> 
Each Openinfra function is defined and described below. 
<br><br>

#### Function: `oi_recode_road_class` 
Below we can see the `recode_road_class` function. This function takes a SimpleFeatures ([sf](https://r-spatial.github.io/sf/index.html)) osm infrastructure network and re-classifies features based
on road classifications specified by [Chan and Cooper](https://www.nature.com/articles/s41598-019-55669-8) in the following [table](https://www.nature.com/articles/s41598-019-55669-8/tables/6).
<br><br>
[Chan and Cooper's](https://www.nature.com/articles/s41598-019-55669-8) work, *"Using road class as a replacement for predicted motorized traffic flow in spatial network models of cycling"*.
<br><br>
Cycling models require the simulation of motorised traffic in order to simulate motorised traffic movement to be considered in cycling models for route planning (cyclists won't enjoy being on a busy road full of traffic!).
<br><br>
To model a given area for cycling, simulated motorised traffic must be modelled for a circular area (∝r^2^) 20x that of the cyclists, as zonal Origin Destination (OD) data is used for the entire car trip (typical car journey range ~30km to a cyclists ~7km).
<br><br>
Accurately modelling motor traffic for such a larger area requires much more data, time, and money than that required to model for cyclists, this becomes even more of an issue when comparing budgets for cyclist modelling and motorised traffic modelling.
<br><br>
[Work](https://www.nature.com/articles/s41598-019-55669-8) by Chan and Cooper propose estimating road traffic based on road classes, as defined in the table above, negating the requirement for the costly and time consuming modelling of motorised traffic.

```{r function, class.source = 'fold-hide'}
# Define recode_road_class function
oi_recode_road_class <- function(osm_sf) {
  # browser() Uncomment this to perform function debugging 
  
  # Created road_class columns
  osm_recat = osm_sf %>%
    # Creates road_class column
    dplyr::mutate(oi_road_class = dplyr::case_when(
      # (7) - Motorways
      highway %in% c("motorway", "motorway_link") ~ "7",
      # (6/5) - Dual Carriageways resi & non-resi
      highway %in% c("trunk", "trunk_link") ~ "6/5",
      # (4) - Primary Roads
      highway %in% c("primary", "primary_link") | (highway == "trunk" & oneway == "F") ~ "4",
      # (3) - Secondary Roads
      highway %in% c("secondary", "secondary_link") ~ "3",
      # (2) - Tertiary Roads
      highway %in% c("tertiary", "tertiary_link") ~ "2",
      # (1) - Local Roads
      highway %in% c("living_street", "residential", "unclassified") ~ "1",
      # (0) - Traffic-free Paths
      highway == "cycleway" ~ "0"
    )) %>%
    
    # Creates road_description columns
    dplyr::mutate(oi_road_desc = dplyr::case_when(
      # (7) - Motorways
      highway %in% c("motorway", "motorway_link") ~ "Motorways",
      # (6/5) - Dual Carriageways resi & non-resi
      (highway %in% c("trunk", "trunk_link")) & (oneway %in% c("yes", "-1", "reversible", "alternating")) ~ "Dual Carriageways (R&NR)",
      # (4) - Primary Roads
      highway %in% c("primary", "primary_link") | (highway %in% c("trunk", "trunk_link")) & (! oneway %in% c("yes", "-1", "reversible", "alternating")) ~ "Primary Roads",
      # (3) - Secondary Roads
      highway %in% c("secondary", "secondary_link") ~ "Secondary Roads",
      # (2) - Tertiary Roads
      highway %in% c("tertiary", "tertiary_link") ~ "Tertiary Roads",
      # (1) - Local Roads
      highway %in% c("living_street", "residential", "unclassified") ~ "Residential / Local Roads",
      # (0) - Traffic-free Paths
      highway == "cycleway" ~ "Cycleway"
    )) %>%
    
    # Removes features that have not been recodeed to a road_class value
    dplyr::filter(!is.na(oi_road_class))
}
```

<br><br>

As discussed, the above function was defined defined based on the following table from Chan and Cooper: 

Road Class | Description | Selection from OSM
-- | -- | --
7 | Motorways | `highway = motorway` **OR** `highway = motorway_link`
6 | Non-residential Dual Carriageways | `highway =  trunk` **OR** `highway = trunk_link` **AND** `oneway = yes`*& manual classification needed*
5 | Residential Dual Carriageways | `highway =  trunk` **OR** `highway =  trunk_link` **AND** `oneway = yes` *& manual classification needed*
4 | Primary Roads | (`highway =  primary` **OR** `highway = primary_link` **OR** `highway = trunk`) **AND** `oneway != yes`)
3 | Secondary Roads | `highway =  secondary` **OR** `highway =  secondary_link`
2 | Tertiary Roads | `highway = tertiary` **OR** `highway = tertiary_link`
1 | Local Roads | `highway = living_street` **OR** `highway = residential` **OR** `highway = unclassified`
0 | Traffic-free Paths | `highway = cycleway`
<br><br>

#### Function: `oi_active_cycle`

Adds column `oi_active_cycle` with values `"yes"|"no"` depending on whether or not a way is cyclable by cyclists. 

```{r function2}
oi_active_cycle = function(osm_sf, remove=FALSE){
  #browser() # Uncomment to debug
  osm_sf_cycle = osm_sf %>% dplyr::mutate(oi_cycle = dplyr::case_when(
    
    # Highway cannot be NA
    is.na(highway) ~ "no",
    
    # Highway values not appropriate for cycling
    highway %in% c('abandoned', 'bus_guideway', 'byway', 'construction', 
                   'corridor', 'elevator', 'fixme', 'escalator', 'gallop', 
                   'historic', 'no', 'planned', 'platform', 'proposed', 
                   'raceway', 'steps') 
    ~ "no",
    
    # Create two cases to catch features typically not allowed for cyclists, but have been flagged appropriate:
    # 1 - If highway = "bad highway values" BUT bicycle = "good bicycle values" then assign ~ "yes",
    (highway %in% c('motorway', 'motorway_link', 'footway', 'bridleway','pedestrian') & bicycle %in% c('yes', 'designated', 'permissive', 'destination')) ~ "yes",
    ((highway == "footway") & (! bicycle %in% c("no", "dismount", "private"))) & (! access %in% c("private", "no", "permit")) ~ "yes",
    
    # 2 - Assign highway = "bad highway values" ONLY as ~ "no". Ways that are appropriate for cyclists will already be assigned "yes" from above
    highway %in% c('motorway', 'motorway_link', 'footway', 'bridleway','pedestrian') ~ "no",
    
    # Way must have access rights
    access %in% c('private', 'no') ~ "no",
    
    # Way must not bar cyclists
    bicycle %in% c('no', 'private', 'ue_sidepath', 'restricted') ~ "no", 
    
    # Way must not contain "private" within the service tag
    grepl("private", service) ~ "no",

  ))
  
  # Case_when above should have added "no" to all inappropriate features, now 
  # find features that do not contain "no" and set as "yes"
  osm_sf_cycle$oi_cycle[is.na(osm_sf_cycle$oi_cycle)] = "yes"
  
  # If remove = TRUE, filter out features that have oi_cycle == "no"
  if (remove){
    osm_sf_cycle = osm_sf_cycle %>% dplyr::filter(osm_sf_cycle$oi_cycle == "yes")
  }
  return(osm_sf_cycle)
}
```
<br><br>

#### Function: `oi_active_walk`

Adds the column `oi_active_walk` with values `"yes"|"no"` depending on whether or not a way is walkable by pedestrians or not.

```{r}
oi_active_walk = function(osm_sf, remove = FALSE){
  #browser() #<-- Uncomment to debug function.
  osm_sf_walking = osm_sf %>% dplyr::mutate(oi_walk = dplyr::case_when(
    
    # Highway tag cannot be NA
    is.na(highway) ~ "no",
    
    # Highway tag cannot have un-walkable values
    highway %in% c('abandoned', 'bus_guideway', 'byway', 'construction', 
                   'corridor', 'elevator', 'fixme', 'escalator', 'gallop', 
                   'historic', 'no', 'planned', 'platform', 'proposed',
                   'raceway', 'motorway', 'motorway_link') 
    ~ "no",
    
    # Hihgway tag cannot be a cycleway UNLESS walking is permitted
    ((highway == "cycleway") & (foot %in% c('yes', 'designated', 'permissive', 'destination'))) ~ "yes",
    # Below catches highway == cycleway and foot != values below INCLUDING NAs. 
    ((highway == "cycleway") & (! foot %in% c('yes', 'designated', 'permissive', 'destination'))) ~ "no",
    
    (highway %in% c("footway", "pedestrian", "path")) & (! foot %in% c("no", "private")) & (! access %in% c("private", "no")) ~ "yes",
    
    # Access cannot be restricted
    ((access %in% c('private', 'no')) & (foot == "yes")) ~ "yes",
    ((access %in% c('private', 'no'))) ~ "no",
    
    # Foot usage must be permitted,
    foot %in% c('private', 'no', 'use_sidepath', 'restricted') ~ "no",
    
    # Service value does not contain "private"
    ((grepl("private", service)) & (foot == "yes")) ~ "yes",
    ((grepl("private", service)) & (foot != "yes")) ~ "no",
    
  ))
  
  # find features that do not contain "no" and set as "yes"
  osm_sf_walking$oi_walk[is.na(osm_sf_walking$oi_walk)] = "yes"
  
  # If remove = TRUE, filter out features that have oi_walking == "no"
  if (remove){
    osm_sf_walking = osm_sf_walking %>% dplyr::filter(osm_sf_walking$oi_walk == "yes")
  }
  
  return(osm_sf_walking)
}
```
<br><br>

#### Function: `oi_clean_maxspeed_uk`

Adds column `oi_maxspeed` that contains cleaned maxspeed values from OSM features, where the maxspeed field exists. 

```{r}
oi_clean_maxspeed_uk = function(osm_sf, no_NA = FALSE, del = FALSE) {
  
  # Define NOT IN 
  `%!in%` = Negate(`%in%`)
  
  osm_clean = osm_sf %>%
    dplyr::mutate(oi_maxspeed = dplyr::case_when(
      # maxspeed == national, when on motorway
      (maxspeed == "national" & highway %in% c("motorway", "motorway_link")) ~ "70 mph",
      
      # maxspeed == national, when NOT on motorway
      (maxspeed == "national" & highway %!in% c("motorway", "motorway_link")) ~ "60 mph",  
      
      # maxspeed == national, when on standard (i.e Non-Residential) dual carriageway
      (maxspeed == "national" & highway %in% c("trunk", "trunk_link")) ~ "70 mph",
      
      # maxspeed == (20|30|40|50|60|70 --> + mph)
      maxspeed == "20" ~ "20 mph",
      maxspeed == "30" ~ "30 mph",
      maxspeed == "40" ~ "40 mph",
      maxspeed == "50" ~ "50 mph",
      maxspeed == "60" ~ "60 mph",
      maxspeed == "70" ~ "70 mph",
      
      # Already cleaned speeds (to stop mutate missing these)
      maxspeed == "20 mph" ~ "20 mph",
      maxspeed == "30 mph" ~ "30 mph",
      maxspeed == "40 mph" ~ "40 mph",
      maxspeed == "50 mph" ~ "50 mph",
      maxspeed == "60 mph" ~ "60 mph",
      maxspeed == "70 mph" ~ "70 mph",
    )) 
  
  if (no_NA){
    # if TURE, will remove features if their oi_maxspeed == NA
    osm_clean = osm_clean %>% dplyr::filter(!is.na(oi_maxspeed))
  }
  
  if (del){
    # If TRUE, will delete original `maxspeed` column
    osm_clean = subset(osm_clean, select = -c(maxspeed))
  } 

  return(osm_clean)
  }

```
<br><br>

#### Function: `oi_inclusive_mobility`

Adds a number of Inclusive Mobility (IM) `"im_***"` columns that reflect whether or not a piece of infrastructure meets the requirements of the [IM Guide](https://www.gov.uk/government/publications/inclusive-mobility-making-transport-accessible-for-passengers-and-pedestrians)

```{r}
oi_inclusive_mobility = function(osm_sf) {

  #browser() # <-- Uncomment for debugging.
  
  osm_sf_im = osm_sf %>% 
    # Assesses whether a kerb is flush or not
    dplyr::mutate(im_kerb = dplyr::if_else(kerb == "flush" | kerb == "no", "flush", "other")) %>% 
    
    # Assesses footway - a ‘pavement’ adjacent to a road
    dplyr::mutate(im_footway = dplyr::case_when(
      footway %in% c("left", "right", "both", "sidewalk") |
        sidewalk %in% c("left", "right", "both", "yes", "separate") |
        # trying to capture footways shared with cyclists
        !is.na(cycleway) & # map cycling infrastructure that is an inherent part of the road
        foot %in% c("yes", "designated") |
        segregated %in% "yes"
      ~ "yes",
      TRUE ~ "no" 
    ) 
    ) %>% 
    # Assesses footpath - any other right of way for pedestrians, that does not run adjacent to a road.
    dplyr::mutate(im_footpath = dplyr::case_when(
      highway %in% "footway" & 
        im_footway %in% "no" | 
        # not (always) an inherent part of the road
        highway %in% c("cycleway", "bridleway", "path") & # foot = "designated" is implied
        im_footway %in% "no" &
        ! foot %in% c("no", "private") | 
        ! access %in% c("no", "private") &
        segregated %in% "no" # shared space
      ~ "yes",
      TRUE ~ "no"
    )
    ) %>%
    
    # Assesses presence of a crossing and what type: give-way, signal controlled, none, or yes (but the type is unknown)
    dplyr::mutate(im_crossing = dplyr::case_when(
      stringr::str_detect(crossing, "zebra|uncontr|marked")~ "give-way",
      stringr::str_detect(crossing, "toucan|pedex|puffin|equestrian|light|signal")~ "signal-controlled",
      highway %in% "crossing" | footway  %in% "crossing" | !is.na(crossing) ~ "yes",
      TRUE ~ "no"
    )) %>% 
    
    # implied footways but there's a lack of data to verify
    dplyr::mutate(im_footway_imp = dplyr::case_when(
      im_footway %in% "no" &
        im_footpath %in% "no" &
        im_crossing %in% "no"
      ~ "yes",
      TRUE ~ "no"
    )
    ) %>% 
    # Assesses whether the way is lit or not
    dplyr::mutate(im_light = dplyr::case_when( 
      # highway %in% "street_lamp" |
      ! lit %in% c("no", "disused") & ! is.na(lit)
      ~ "yes",
      TRUE ~ "no"
    )
    ) %>% 
    
    # Assesses the presence of tactile paving - either yes, no.
    dplyr::mutate(im_tactile = dplyr::case_when(
      ! tactile_paving %in% c("no", "incorrect", "bad") & ! is.na(tactile_paving) 
      ~ "yes",
      ! is.na(tactile_paving)
      ~ "no"
    )
    ) %>%
    
    # Assesses whether surface is paved, unpaved, or other
    dplyr::mutate(
      im_surface_paved = dplyr::case_when(
        highway %in% "cycleway"
        ~ "paved",
        
        stringr::str_detect(surface,
                            "pav|asph|chipseal|concrete|paving|sett|cobble|metal|wood|stepping")
        ~ "paved",
        highway %in% c("footway", "bridleway") & # highway = footway implied surface value is unpaved
          ! surface %in% stringr::str_detect(surface, "pav|asph|chipseal|concrete|paving|sett|cobble|metal|wood|stepping")
        ~ "unpaved",
        stringr::str_detect(surface, "unpav|compact|gravel|rock|pebble|ground|dirt|grass|mud|sand|woodchips|snow|ice|salt")
        ~ "unpaved",
        TRUE & !is.na(surface) ~ "other"
      )
    ) %>% 
    # Assesses whether surface is even or uneven
    dplyr::mutate(im_surface = dplyr::case_when(
      stringr::str_detect(surface, "asph|concrete")
      ~ "even",
      
      im_surface_paved %in% "paved" &
        smoothness %in% c("excellent", "good")
      ~ "even",
      ! is.na(im_surface_paved) 
      ~ "uneven"
    )
    ) %>% 
    # Assesses way width - either under 1.5 meters, 1.5-2 meters, or over 2 meters
    dplyr::mutate(
      im_width =  width %>% 
        readr::parse_number(),
      im_width = dplyr::case_when(
        im_width > 0 & im_width < 1.5 ~ " < 1.5",
        im_width <= 1.5 & im_width <= 2 ~ "1.5 - 2",
        im_width > 2 ~ "> 2"
      )
    ) %>% 
    # Assesses estimated way width - either under 1.5 meters, 1.5-2 meters, or over 2 meters
    dplyr::mutate(
      im_width_est = est_width %>% 
        readr::parse_number(),
      im_width_est = dplyr::case_when(
        im_width_est > 0 & im_width_est < 1.5 ~ "< 1.5",
        im_width_est <= 1.5 & im_width_est <= 2 ~ "1.5 - 2",
        im_width_est > 2 ~ "> 2"
      )
    )
  return(osm_sf_im)
}
```
<br><br>

#### Function: `oi_is_lit`

Adds column `oi_is_lit` with values `"yes"|"no"|"yes- LED"|"no -LED"|"maybe"` depending on whether the OSM way is lit or not, and if the lighting is LED or not. The value `"maybe"` implies that there is a lack of data to be able to distinguish - this is likely "yes" in built up urban areas, and "no" for more rural areas.

```{r}
oi_is_lit = function(osm_sf, remove=FALSE){
  
  # Select only the ways that have a highway value (not train tracks etc.)
  osm_sf = osm_sf %>% dplyr::filter(! is.na(osm_sf$highway))
  
  # Recategorise the data
  osm_sf_lit = osm_sf %>% dplyr::mutate(oi_is_lit = dplyr::case_when( 
    # If lit=* is not a unlit value - set as yes.
    (! lit %in% c("no", "disused") & ! is.na(lit)) ~ "yes",
    # If way is known to be unlit
    lit %in% c("no", "disued") ~ "no",
    
    # Used if street light is an LED (increasing popularity)
    lit_by_led == "yes" ~ "yes - LED",
    lit_by_led == "no" ~ "no - LED",
    ))
  
  # Finally - all remaining N/A are unknown as either lit or unlit.
  osm_sf_lit$oi_is_lit[is.na(osm_sf_lit$oi_is_lit)] = "maybe"
  
  # If remove = TRUE, filter out features that have oi_is_lit == "no"
  if (remove){
    osm_sf_lit = osm_sf_lit %>% dplyr::filter(osm_sf_lit$oi_is_lit %in% c("yes", "yes - LED", "maybe"))
  }
  
  return(osm_sf_lit)
}
```


<br><br>
<br><br>

## Example Data Pack Production - Leeds

Here we go over a reproducible example of generating a transport infrastructure data pack from a OSM network retrieved using [`osmextract`](https://github.com/ropensci/osmextract). 


<br><br>
This example concerns the LAD of Leeds, UK. <br>
Note that a 7.5km circular buffer has been applied (centered at Long, Lat coords (-1.548567, 53.801277)) to reduce map filesize so that the visualisations below can be hosted on this webpage.
<br><br>

### Getting Data
```{r download_data}
# Data downloaded through osmextract as shown below: 

#total_place = osmextract::oe_get(
#  place = "Leeds",
#  provider = "bbbike",
#  layer = "lines",
#  never_skip_vectortranslate = TRUE,
#  force_download = TRUE,
#  quiet = FALSE,
#  extra_tags = c("foot", "bicycle", "access", "service", "maxspeed", "oneway",
#                  "kerb", "footway", "sidewalk", "cycleway", "segregated", "highway",
#                  "crossing", "lit", "tactile_paving", "surface", "smoothness", "width",
#                  "est_width", "lit_by_led", "boundary", "admin_level", "name")
#)

# For reproduciblity you may chose to use the example dataset from our releases.
# You can create your own data using the oe_get() function above.
leeds_network = sf::read_sf("https://github.com/udsleeds/openinfra/releases/download/v0.2/Leeds.geojson")
```

<br><br>

### Applying Openinfra Functions - Adding Value 
```{r function_application, eval=FALSE}

leeds_network = oi_recode_road_class(leeds_network) 
leeds_network = oi_active_walk(leeds_network)
leeds_network = oi_active_cycle(leeds_network)
leeds_network = oi_clean_maxspeed_uk(leeds_network)
leeds_network = oi_inclusive_mobility(leeds_network)
leeds_network = oi_is_lit(leeds_network)
```
<br><br>

### Selecting Relevant Columns for Data Pack
```{r, create_data_pack, eval=FALSE}
leeds_datapack = leeds_network %>% dplyr::select(c(
 "osm_id", "highway", "oi_road_desc", "oi_maxspeed", "oi_walk", "oi_cycle",
 "oi_is_lit", "im_kerb", "im_footway", "im_footpath", "im_crossing", 
 "im_footway_imp", "im_light", "im_tactile", "im_surface_paved", "im_surface",
 "im_width", "im_width_est"))
```
Thus concludes the creation of the transport infrastructure data packs.
<br><br>
We can visualise these outputs below.

<br><br><br><br>

