---
title: "Getting started with open data on transport infrastructure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with open data on transport infrastructure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# What is OpenStreetMap?

OpenStreetMap (OSM) has contributed to the shift in perception of who can map and how it can be done (Haklay and Weber 2008). Conceived in 2004 by Steve Coast, the project aims to create a free and editable map that everyone could access and use. The project's focus on making its data accessible to local communities is one of the reasons behind OSM being community-driven with an emphasis on local knowledge in mapping. Putting community and local knowledge at the heart has many benefits, such as: 

- keeping data up-to-date
- empowering citizens to represent their local environments
- encouraging the participation of citizens in policy making

The importance of local knowledge is also highlighted in the LTN 1/20 guide in the context of ensuring successful implementation of a scheme. Hence, OSM can be a tool to foster a bottom-up approach in active travel planning through the inclusion of citizens in the data generation process. This, consequently, can be used to vocalise their needs leading to an increased likelihood of achieving higher uptake levels.

# What is the structure of OpenStreetMap data?

The structure of OSM data is composed of three elements, or entities: 

- nodes (points defined by latitude and longitude)
- ways (linestrings that can also be used to represent polygons)
- relations (define relationships between two or more data elements (e.g., nodes and ways))

## Nodes
```{r, eval = FALSE, echo = FALSE}
q = "SELECT * FROM 'points' WHERE amenity IN ('bicycle_rental')"
extra_tags = "amenity"
region_name = "Leeds"
bicycle_parking_points = oe_get(place = region_name, 
                               query = q, 
                               extra_tags = extra_tags,
                               layer = "points",
                               provider = "bbbike",
                               force_vectortranslate = TRUE,
                               force_download = TRUE)
bicycle_parking_points %>% saveRDS("bicyle_parking_points.Rds")
```

```{r, out.width="100%", echo=FALSE, message=FALSE, warning=FALSE}
url = "http://github.com/udsleeds/openinfra/releases/download/v0.1/bicyle_parking_points.Rds"
download.file(url = url,
              destfile = "bicyle_parking_points.Rds")

bicycle_parking_points = readRDS("bicyle_parking_points.Rds")
tmap::tmap_mode("view")
tmap::qtm(bicycle_parking_points,
    dots.col = "red"
    )
```

