---
title: "Getting started with open data on transport infrastructure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with open data on transport infrastructure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# What is OpenStreetMap?

OpenStreetMap (OSM) has contributed to the shift in perception of who can map and how it can be done (Haklay and Weber 2008). Conceived in 2004 by Steve Coast, the project aims to create a free and editable map that everyone could access and use. The project's focus on making its data accessible to local communities is one of the reasons behind OSM being community-driven with an emphasis on local knowledge in mapping. Putting community and local knowledge at the heart has many benefits, such as: 

- keeping data up-to-date
- empowering citizens to represent their local environments
- encouraging the participation of citizens in policy making

The importance of local knowledge is also highlighted in the LTN 1/20 guide in the context of ensuring successful implementation of a scheme. Hence, OSM can be a tool to foster a bottom-up approach in active travel planning through the inclusion of citizens in the data generation process. This, consequently, can be used to vocalise their needs leading to an increased likelihood of achieving higher uptake levels. 

# What is the structure of OpenStreetMap data?

**note that OSM has not strict protocols in place**

The structure of OSM data is composed of three elements, or entities: 

- nodes 
- ways 
- relations 

In general, OSM can be described as a graph with edges and vertices that are connected or isolated (Bennett 2010). Tags (key="value" pairs) are used to describe their geographical features.

## Nodes

Nodes are points in space that are defined by latitude and longitude. They can be used on their own or as part of a way or a relation.

An example of a point in active transport context would be a bicycle rental.

![Bicycle rental in OSM in-browser editor ([link](https://www.openstreetmap.org/edit#map=19/53.75200/-1.55801))
](/Users/gretatimaite/Desktop/openinfra/figures/bicycle_rental_node_edited.jpg)
```{r, eval = FALSE, echo = FALSE}
q = "SELECT * FROM 'points' WHERE amenity IN ('bicycle_rental')"
extra_tags = "amenity"
region_name = "Leeds"
bicycle_parking_points = oe_get(place = region_name, 
                               query = q, 
                               extra_tags = extra_tags,
                               layer = "points",
                               provider = "bbbike",
                               force_vectortranslate = TRUE,
                               force_download = TRUE)
bicycle_parking_points %>% saveRDS("bicyle_parking_points.Rds")
```

```{r, title = "bicycle rental", out.width="100%", echo=FALSE, message=FALSE, warning=FALSE}
url = "http://github.com/udsleeds/openinfra/releases/download/v0.1/bicyle_parking_points.Rds"
download.file(url = url,
              destfile = "bicyle_parking_points.Rds")

bicycle_parking_points = readRDS("bicyle_parking_points.Rds")
tmap::tmap_mode("view")
tmap::qtm(bicycle_parking_points,
    dots.col = "red"
    )
```

## Ways

Ways are lines formed by linking at least two nodes (i.e., representing two ends), however the node can belong to more than one way. They always have a direction, represented by the arrows in the editor, even if it is not meaningful (e.g., a wall). These kinds of ways can be defined as open ways as they have different start and end nodes.

Ways can also be closed if the start and end nodes are identical. This is usually used to represent areas, or polygons, but also roundabouts or closed barriers, such as a closed wall. 

An example of a way is a cycleway in a park next to the bicycle rental.
![Cycleway in OSM in-browser editor ([link](https://www.openstreetmap.org/edit#map=17/53.75158/-1.55801))
](/Users/gretatimaite/Desktop/openinfra/figures/way_cycleway_edited.jpg)

A park itself is a good example of an area.
![Park in OSM in-browser editor ([link](https://www.openstreetmap.org/edit#map=14/53.7587/-1.5471))
](/Users/gretatimaite/Desktop/openinfra/figures/park_area_edited.jpg)

## Relations

Relation is a list of nodes, ways, and/or relations. It is used to model geographic relationships between objects. These relations should not contain only closely connected elements, having no more than 300 members (elements) per relation. 

Most likely the most important relation in the context of active transport is route but there are also other relations, such as public transport schemes and restrictions. 

![Way as a member of a relation forming a cycleway route ([link](https://www.openstreetmap.org/edit#map=16/53.7568/-1.5410))
](/Users/gretatimaite/Desktop/openinfra/figures/cycleway_relation_edited.jpg)


## Tags

Tags are not *exactly* data structures but they play a vital role in OSM because they define geographic features. In other words, it would be without tags one would not know if a node represents a bicycle rental.

A tag is a concept that refers to a key="value" pair. A key describes a topic while value is a specification of a key. For instance, highway="cycleway" indicates that it is a highway (i.e., a line) and, specifically, a cycleway. It is important to note that a value is not always unique in a sense that it can also be a key. For example, there is a "cycleway" tag, which is used to further detail cycling infrastructure. Hence, it is important to pay attention to the syntax of a tag to understand its meaning and use in a given context.

Moreover, each OSM element can have multiple tags and the use of tags is not limited to certain data elements. Thus, theoretically, a point could be `highway="footway"`. The lack of restrictions is inherent in OSM. It relies on the mappers to make decisions that are the most sensible in their specific contexts. This does not mean, however, that there are no conventions about the most appropriate way of tagging something. A simple example would be a depreciation of a sidewalk="none" tag which should be changed to sidewalk="no". A more intricate example is a question concerning the mapping of sidewalks as currently there are two proposed schemes: one considering a sidewalk as a tag on a highway and another proposing to map footways as separate ways. 
The table below outlines some relevant tags for active travel planning that should be intuitive enough to get started.



A list of links:

- https://wiki.openstreetmap.org/wiki/Types_of_relation
- https://learnosm.org/en/josm/josm-relations/
- https://wiki.openstreetmap.org/wiki/Elements
- https://labs.mapbox.com/mapping/osm-data-model/
- https://wiki.openstreetmap.org/wiki/Forward_%26_backward,_left_%26_right#Identifying_the_direction_of_a_.27way.27










