---
title: "Getting started with open data on transport infrastructure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with open data on transport infrastructure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# What is OpenStreetMap?

OpenStreetMap (OSM) has contributed to the shift in perception of who can map and how it can be done (Haklay and Weber 2008). Conceived in 2004 by Steve Coast, the project aims to create a free and editable map that everyone could access and use. The project's focus on making its data accessible to local communities is one of the reasons behind OSM being community-driven with an emphasis on local knowledge in mapping. Putting community and local knowledge at the heart has many benefits, such as: 

- keeping data up-to-date
- empowering citizens to represent their local environments
- encouraging the participation of citizens in policy making

The importance of local knowledge is also highlighted in the LTN 1/20 guide in the context of ensuring successful implementation of a scheme. Hence, OSM can be a tool to foster a bottom-up approach in active travel planning through the inclusion of citizens in the data generation process. This, consequently, can be used to vocalise their needs leading to an increased likelihood of achieving higher uptake levels. 

# What is the structure of OpenStreetMap data?

OSM relies on a simple data structure and flexible tagging to describe geographical objects. Indeed, OSM deliberately chose not to adapt the existing standards for geographic information (Halklay and Weber, 2008) as the goal was to simplify the use and maintenance of a project. 

The structure of OSM data is composed of three key elements (also called entities or primitives):

- nodes 
- ways 
- relations 

In general, OSM can be described as a graph with edges and vertices that are connected or isolated (Bennett 2010). Tags (key="value" pairs) are used to describe their geographical features.

## Nodes

Nodes are points in space that are defined by latitude and longitude. They can be used on their own or as part of a way or a relation.

An example of a point in active transport context would be a bicycle rental.

```{r, echo=FALSE}
url_image1 = "https://github.com/udsleeds/openinfra/blob/main/figures/bicycle_rental_node_edited.jpg?raw=true"
```
![image](`r url_image1`)

## Ways

Ways are lines formed by linking at least two nodes (i.e., representing two ends), however the node can belong to more than one way. They always have a direction, represented by the arrows in the editor, even if it is not meaningful (e.g., a wall). These kinds of ways can be defined as open ways as they have different start and end nodes.

Ways can also be closed if the start and end nodes are identical. This is usually used to represent areas, or polygons, but also roundabouts or closed barriers, such as a closed wall. 

```{r, echo=FALSE}
url_image2 = "https://github.com/udsleeds/openinfra/blob/main/figures/way_cycleway_edited.jpg?raw=true"
```
![image](`r url_image2`)

A park itself is a good example of an area.
```{r, echo=FALSE}
url_image3 = "https://github.com/udsleeds/openinfra/blob/main/figures/park_area_edited.jpg?raw=true"
```
![image](`r url_image3`)


## Relations

Relation is a list of nodes, ways, and/or relations. It is used to model geographic relationships between objects. These relations should not contain only closely connected elements, having no more than 300 members (elements) per relation. 

Most likely the most important relation in the context of active transport is route but there are also other relations, such as public transport schemes and restrictions. 

```{r, echo=FALSE}
url_image4 = "https://github.com/udsleeds/openinfra/blob/main/figures/cycleway_relation_edited.jpg?raw=true"
```
![image](`r url_image4`)

## Tags

Tags are not *exactly* data structures but they play a vital role in OSM because they define geographic features. In other words, without tags one would not know if a node represents a bicycle rental.

A tag is a concept that refers to a key="value" pair. A key describes a topic while value is a specification of a key. For instance, highway="cycleway" indicates that it is a highway (i.e., a line) and, specifically, a cycleway. It is important to note that a value is not always unique in a sense that it can also be a key. For example, there is a "cycleway" tag, which is used to further detail cycling infrastructure. Hence, it is important to pay attention to the syntax of a tag to understand its meaning and use in a given context.

Moreover, each OSM element can have multiple tags and the use of tags is not limited to certain data elements. Thus, theoretically, a point could be `highway="footway"`. The lack of restrictions is inherent in OSM. It relies on the mappers to make decisions that are the most sensible in their specific contexts. This does not mean, however, that there are no conventions about the most appropriate way of tagging something. A simple example would be a depreciation of a sidewalk="none" tag which should be changed to sidewalk="no". A more intricate example is a question concerning the mapping of sidewalks as currently there are two proposed schemes: one considering a sidewalk as a tag on a highway and another proposing to map footways as separate ways. 
The table below outlines some relevant tags for active travel planning that should be intuitive enough to get started.

```{r, results="hide", echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(tmap)

url = "https://github.com/udsleeds/openinfra/releases/download/v0.1/leeds_central_15-02-2022.RDS"
download.file(url,
              destfile = "leeds_central_15-02-2022.RDS")
leeds_central = sf::read_sf("leeds_central_15-02-2022.RDS") # reading the file

```

```{r, echo=FALSE, message=FALSE, warning=FALSE, out.width="100%"}
tmap::tmap_mode("view")

cl_highways = leeds_central %>%
  dplyr::select(highway)
cl_cycleways = leeds_central %>%
  dplyr::select(cycleway)
cl_footways = leeds_central %>%
  dplyr::select(footway)
cl_maxspeed = leeds_central %>% 
  dplyr::select(maxspeed)
cl_kerb = leeds_central %>% 
  dplyr::select(kerb)
cl_wheelchair = leeds_central %>% 
  dplyr::select(wheelchair)
cl_width = leeds_central %>% 
  dplyr::select(width)
cl_lit = leeds_central %>% 
  dplyr::select(lit)

tmap::tm_shape(cl_highways)+
  tmap::tm_lines("highway")+
  tmap::tm_shape(cl_cycleways)+
  tmap::tm_lines("cycleway")+
  tmap::tm_shape(cl_footways)+
  tmap::tm_lines("footway")+
  tmap::tm_shape(cl_maxspeed)+
  tmap::tm_lines("maxspeed")+
  tmap::tm_shape(cl_kerb)+
  tmap::tm_lines("kerb")+
  tmap::tm_shape(cl_wheelchair)+
  tmap::tm_lines("wheelchair")+
  tmap::tm_shape(cl_width)+
  tmap::tm_lines("width")+
  tmap::tm_shape(cl_lit)+
  tmap::tm_lines("lit")
```
Important: No highways in central Leeds have width tagged.


# OSM and transport research

In 2017, Barrington-Leigh and (**reference**) argued that over 80% of all roads are mapped in OSM and most of the European and North American countries reaching more than 95%. While the extensiveness of the road network is impressive, the paper focused on roads intended for vehicle circulation. While it can be assumed that some of the roads can and are used by, for instance, cyclists but it does not reveal anything about the cycling infrastructure which is needed to promote active modes of travel (Handy et al. 2014). To address this gap, Ferster et al. (**reference**) conducted a study  in which the data of OSM was compared to official open data from city government in 6 Canadian cities. They found that there was a high level of concordance between the datasets, but that consistent labelling (i.e., topology) of cycling infrastructure in OSM is a challenge that needs to be considered. Importantly, however, it is noted that some of the inconsistency might have resulted from the fact that some types of cycling infrastructure is new in Canada, such as cycling tracks. Although it shows that it might take some time for OSM mappers to (re)map their surroundings, it also calls attention to the importance of communicating new infrastructural developments so that different types can be easily identified, mapped, and tagged.

While there seems to be no research similar to that conducted by Ferster et al. (2020) in the context of the UK, there are examples of OSM utilization to study active travel. For example, Orozco et al. (2020) outlined an algorithm to improve cycling network by connecting disconnected segments. They argued that for a highly segmented city like London (over 3000 segments compared to Copenhagen's 300) it will take great investments to improve connectivity. Moreover, OSM has also been joined to other datasets, such as Strava or GPS and travel diary data to improve the understanding of cyclists route choice in Glasgow (Alattar et al. 2021) and North England (Yeboah and Alvanides 2015), respectively.

OSM was also used in the context of pedestrian network planning. For example, Kasemsuppakorn and Karimi (2013) examined OSM for pedestrian network construction, Graser (2016) used OSM to evaluate different approaches to integrating open spaces in pedestrian routing systems while Novack et al. (2018) developed a customized pleasant pedestrian routing system. Differently from cycling, however, pedestrian networks appear to be examined more in relation to accessibility. Recently Cohen and Dalyot used OSM to develop a wayfinding algorithm for planning accessible and safe routes to blind pedestrians. Boularouk et al. (2017) argued that the use of OSM could help to develop cheaper assistive technologies for people with disabilities who, compared to people without disabilites, have lower median incomes in the UK (Francis-Devine 2021). However, neither of the mentioned projects have been done in the context of the UK. It seems that, at least in the academic research, there is more attention paid to the use of OSM for cycling network planning compared to walking or wheeling.

## Why use OSM?

So far, we have outlined what is OSM and how it was used in transport research. However, what kind of benefits would the adoption of open data yield? 

The first one, already mentioned, is the potential of OSM to reduce costs of the tools and systems that rely on geospatial data (Bouldarouk et al. 2017). Second, OSM could be used for conducting comparative analysis between different countries and/or regions, especially in the absence of national geographic dataset (**reference**). Third, the use of OSM data could increase the transparency, reproducibility, and replicability of (geospatial) research and, thus, its robustness (Brunsdon and Comber 2020). Moreover, opening spatial analyses in general, and the data used in particular, would unveil the "black box" algorithms that, in the age of Big Data, constitute a part in the daily lives of citizens (Brunsdon and Comber 2020). Fourth, it fosters citizen science and citizen participation in democratic processes by empowering citizens to engage in (local) geographic knowledge production (see Peters 2019). Given that OSM mappers tend to be highly engaged with their local communities (**reference**), adoption of OSM could lead to increased accountability and promotion of evidence-based decision making in transport research (Lovelace 2021).









A list of links:

- https://wiki.openstreetmap.org/wiki/Types_of_relation
- https://learnosm.org/en/josm/josm-relations/
- https://wiki.openstreetmap.org/wiki/Elements
- https://labs.mapbox.com/mapping/osm-data-model/
- https://wiki.openstreetmap.org/wiki/Forward_%26_backward,_left_%26_right#Identifying_the_direction_of_a_.27way.27










