---
title: "Network Definitions by package"
author: "James Hulse"
date: '2022-06-15'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<br><br>

*This vignette is currently a work in progress and changes are likely (20/6/22)* <br><br>
This Vignette looks at the pyrosm, OSMnx, and osmsxtract packages for obtaining network data from OpenStreetMaps. 
<br><br>
Here, we look at how each package defines one of three pre-defined networks: 

- Driving Network
- Walking Network 
- Cycling Network

<br>
With OpenInfra's specific focus on active travel and infrastructure our focus is on Cycling and Walking networks. <br>
Driving networks have been included for completeness. 
<br><br>
Understanding how such packages define networks will allow us to analyse and combine the tag filters used by each package to define, hopefully, the best defined active travel networks. 
<br><br>

## osmextract network filters

<br>
Firstly we go through the osmextract (R) package function documentation to see which filters are applied to each of the three network types.
<br><br>
osmextract creates networks with the following workflow. A place name string (i.e. "Leeds") is matched to one of two dataset providers, [BBBike](https://download.bbbike.org/osm/bbbike/) or [Geofabrik](https://download.geofabrik.de/) before downloading a compressed [`.pbf`](https://wiki.openstreetmap.org/wiki/PBF_Format) file of OSM data for the specified place (Leeds). Next the downloaded file is converted into the open standard [`.gpkg`](https://www.geopackage.org/) format. This file is then parsed with an SQL query to obtain network features specified by the user (i.e. "walking", "cycling", or "driving").
<br><br>
Here we look at the three pre-defined filters offered by `osmextract` and how these are parsed with an SQL query, we then try to translate these queries into a python format, similar to the filtering usilised by pyrosm/OSMnx.
<br><br>
After, we compare the tag filtering applied by each package to obtain our final networks.

### Driving Filter
<br>

##### [osmextract](https://github.com/ropensci/osmextract/blob/master/R/get-network.R#L230) (R & SQL)
```{r }
# A motorcar/motorcycle mode of transport includes the following scenarios:
# - highway IS NOT NULL (since usually that means that's not a road);
# - highway NOT IN ('bus_guideway', 'byway' (not sure what it means), 'construction',
# 'corridor', 'cycleway', 'elevator', 'fixme', 'footway', 'gallop', 'historic',
# 'no', 'pedestrian', 'platform', 'proposed', 'steps', 'pedestrian',
# 'bridleway', 'path', 'platform');
# - access NOT IN ('private', 'no');
# - service NOT ILIKE 'private%';
load_options_driving = function(place) {
  list(
    place = place,
    layer = "lines",
    extra_tags = c("access", "service"),
    vectortranslate_options = c(
    "-where", "
    (highway IS NOT NULL)
    AND
    (highway NOT IN (
    'abandoned', 'bus_guideway', 'byway', 'construction', 'corridor', 'elevator',
    'fixme', 'escalator', 'gallop', 'historic', 'no', 'planned', 'platform',
    'proposed', 'cycleway', 'pedestrian', 'bridleway', 'path', 'footway',
    'steps'
    ))
    AND
    (access NOT IN ('private', 'no'))
    AND
    (service NOT ILIKE 'private%')
    "
    )
  )
}
```
<br><br>

#### readable SQL translation
```
Features must have a valid value for highway tag [i.e. highway!=None] (highway IS NOT NULL)

Features are removed if highway tag = any of the following values:
('abandoned', 'bus_guideway', 'byway', 'construction', 'corridor', 'elevator',
    'fixme', 'escalator', 'gallop', 'historic', 'no', 'planned', 'platform',
    'proposed', 'cycleway', 'pedestrian', 'bridleway', 'path', 'footway',
    'steps')
    
Features are removed if access tag = any of the following values: ("private", "no")

Features are removed if the service tag value (a string) contains the sub-string ("private") 
```
<br><br>

#### python representation
```{python}
osmextract_filter = dict() 
osmextract_filter["driving"] = (
  f'["highway]'
  f'["highway!~"abandoned|bus_guideway|byway|construction|corridor|elevator|fixme|escalator|gallop|historic|no|'
  f'planned|platform|proposed|cycleway|pedestrian|bridleway|path|footway|steps"]'
  f'["access"!~"private|no"]'
  f'["service"!~"private"]' #Not too sure how to represent this
)

# Note that the bottom line (f'["service"!~"private"]') should be any value for the service tag should not contain the pattern "private", no matter the actual whole string
```
<br><br><br>

### Walking Filter

#### [osmextract](https://github.com/ropensci/osmextract/blob/master/R/get-network.R#L197) (R & SQL)
``` {r}
# See also https://wiki.openstreetmap.org/wiki/Key:footway and
# https://wiki.openstreetmap.org/wiki/Key:foot
# A walking mode of transport includes the following scenarios:
# - highway IS NOT NULL (since usually that means that's not a road);
# - highway NOT IN ('abandoned', 'bus_guideway', 'byway', 'construction', 'corridor',
# 'elevator', 'fixme', 'escalator', 'gallop', 'historic', 'no', 'planned',
# 'platform', 'proposed', 'raceway', 'motorway', 'motorway_link');
# - highway != 'cycleway' OR foot IN ('yes', 'designated', 'permissive', 'destination');
# - access NOT IN ('no', 'private');
# - foot NOT IN ('no', 'use_sidepath', 'private', 'restricted')
# - service does not look like 'private' (ILIKE is string matching case insensitive)
load_options_walking = function(place) {
  list(
    place = place,
    layer = "lines",
    extra_tags = c("access", "foot", "service"),
    vectortranslate_options = c(
    "-where", "
    (highway IS NOT NULL)
    AND
    (highway NOT IN ('abandoned', 'bus_guideway', 'byway', 'construction', 'corridor', 'elevator',
    'fixme', 'escalator', 'gallop', 'historic', 'no', 'planned', 'platform', 'proposed', 'raceway',
    'motorway', 'motorway_link'))
    AND
    (highway <> 'cycleway' OR foot IN ('yes', 'designated', 'permissive', 'destination'))
    AND
    (access NOT IN ('private', 'no'))
    AND
    (foot NOT IN ('private', 'no', 'use_sidepath', 'restricted'))
    AND
    (service NOT ILIKE 'private%')
    "
    )
  )
}
```
<br><br>

#### readable SQL translation

```
Features must have a valid value for highway tag [i.e. highway!=None] (highway IS NOT NULL)

Features are removed if they contain highway tag = any of the following values:
('abandoned', 'bus_guideway', 'byway', 'construction', 'corridor', 'elevator',
    'fixme', 'escalator', 'gallop', 'historic', 'no', 'planned', 'platform', 'proposed', 'raceway',
    'motorway', 'motorway_link')
    
Features are removed if they contain highway tag = ("cycleway") UNLESS these features contain the tag:value 
    combination of foot = ("yes", "designated", "permissive", "destination") (i.e this cycleway IS pedestrian friendly too)
    
Features are removed if they contain access tag = ("private", "no")

Features are removed if they contain foot tag = ("private", no", "use_sidepath", "restricted")

Features are removed if the service tag value (a string) contains the sub-string ("private") 
```
<br><br>

#### python representation
```{python}
osmextract_filter = dict()
osmextract_filter["walking"] = (
  f'["highway"]'
  f'["highway"!~"abandoned|bus_guideway|byway|construction|corridor|elevator|fixme|'
  f'escalator|gallop|historic|no|planned|platform|proposed|raceway|motorway|motorway_link"]'
  f'["highway!~"cycleway"] UNLESS ["foot"~"yes|designated|permissive|destination"]' # I know this is not proper python. will update once I know more. 
  f'["access"!~"private|no"]'
  f'["foot"!~"private|no|use_sidepath|restricted"]'
  f'["service"!~"private"]' # Not too sure how to represent this
  
)
```
<br><br><br>

### Cycling Filter

#### [osmextract](https://github.com/ropensci/osmextract/blob/master/R/get-network.R#L157) (R & SQL) 
``` {r}
# A cycling mode of transport includes the following scenarios:
# - highway IS NOT NULL (since usually that means that's not a road);
# - highway NOT IN ('abandoned', 'bus_guideway', 'byway', 'construction', 'corridor',
# 'elevator', 'fixme', 'escalator', 'gallop', 'historic', 'no', 'planned',
# 'platform', 'proposed', 'raceway', 'steps');
# - highway NOT IN IN ('motorway', 'motorway_link', 'bridleway', 'footway',
# 'pedestrian) OR bicycle IN ('yes', 'designated', 'permissive', 'destination');
# - access NOT IN ('no', 'private');
# - bicycle NOT IN ('no', 'use_sidepath')
# - service does not look like 'private' (ILIKE is string matching case insensitive)
load_options_cycling = function(place) {
  list(
    place = place,
    layer = "lines",
    extra_tags = c("access", "bicycle", "service"),
    vectortranslate_options = c(
    "-where", "
    (highway IS NOT NULL)
    AND
    (highway NOT IN (
    'abandoned', 'bus_guideway', 'byway', 'construction', 'corridor', 'elevator',
    'fixme', 'escalator', 'gallop', 'historic', 'no', 'planned', 'platform',
    'proposed', 'raceway', 'steps'
    ))
    AND
    (highway NOT IN ('motorway', 'motorway_link', 'footway', 'bridleway',
    'pedestrian') OR bicycle IN ('yes', 'designated', 'permissive', 'destination')
    )
    AND
    (access NOT IN ('private', 'no'))
    AND
    (bicycle NOT IN ('private', 'no', 'use_sidepath', 'restricted'))
    AND
    (service NOT ILIKE 'private%')
    "
    )
  )
}
```
<br><br>

#### readable SQL translation 
```
Features must have a valid value for highway tag [i.e. highway!=None] (highway IS NOT NULL)

Features are removed if they contain highway tag = any of the following values:
('abandoned', 'bus_guideway', 'byway', 'construction', 'corridor', 'elevator',
    'fixme', 'escalator', 'gallop', 'historic', 'no', 'planned', 'platform',
    'proposed', 'raceway', 'steps')
    
Features are removed if they contain highway tag = ("motorway", "motorway_link", "footway", "bridleway") UNLESS these
    features ALSO contain the tag:value combination of bicycle = ("yes", "designated", "permissive", "destination")
    
Features are removed if they contain the tag access = ("private", "no")

Features are removed if they contain the tag bicycle = ("private", "no", "use_sidepath", "restricted")

Features are removed if the service tag value (a string) contains the sub-string ("private") [service NOT ILIKE 'private%']
```
<br><br>

#### python representation
```{python}
osmextract_filter = dict() 
osmextract_filter["cycling"] = (
  f'["highway"]'
  f'["highway"!~"abandoned|bus_guideway|byway|construction|corridor|elevator|fixme|escalator|gallop'
  f'historic|no|planned|platform|proposed|raceway|steps"]'
  f'["highway"!~"motorway|motorway_link|footway|bridleway|pedestrian] UNLESS ["bicycle"~"yes|designated|permissive|destination"]' # I know this is not correct Python - looking into this (i.e. A unless B)
  f'["access"!~"private|no"]'
  f'["bicycle"!~"private|no|use_sidepath|restricted"]'
  f'["service"!~"private"]' # Not sure how to represent in Python (grepl in R)
)

```
<br><br><br><br>

## General Network Definitions (by package)

The pyrosm package defines networks by first taking a complete network from OSM. It then recursively removes features that contain tags matching specific networks described below: 
<br><br>
Definitions have been obtained from [pyrosm source code](https://github.com/HTenkanen/pyrosm/blob/1d93eb050a37e4564b56f49ef816e409b6ae658c/pyrosm/config/osm_filters.py#L1), the [OSMnx source code](https://github.com/gboeing/osmnx/blob/77fd41c4e8d3f87aa93a23193f3f604c263996f8/osmnx/downloader.py#L28), and the [osmextract source code](https://github.com/ropensci/osmextract/blob/master/R/get-network.R).
<br><br> 
It should be noted that for all OSMnx OSM API queries the "highway" tag is specified.<br>
*specifying way["highway"] means that all ways returned must have a highway tag. the specific filters then
remove ways by tag/value.*
<br><br>
**Cycling Filter** removed tags

| Tag | Pyrosm Values| OSMnx Values | osmextract|Proposed Filter|
|:----:|:----:|:----:|:---:|:---:|
|area| "yes" | "yes"|**None**|"yes"|
|access| **None** | "private"| "private", "no"|"private", "no"|
|highway|"footway", "steps", "corridor", "elevator", "escalator", "motor", "proposed", "construction", "abandoned", "platform", "raceway", "motorway", "motorway_link"| "abandoned", "bus_guideway", "construction", "corridor", "elevator", "escalator", "footway", "motor", "planned", "platform", "proposed", "raceway", "steps"|'abandoned', 'bus_guideway', 'byway', 'construction', 'corridor', 'elevator', 'fixme', 'escalator', 'gallop', 'historic', 'no', 'planned', 'platform', 'proposed', 'raceway', 'steps' |'abandoned', 'bus_guideway', 'byway', 'construction', 'corridor', 'elevator', 'fixme', 'escalator', 'gallop', 'historic', 'no', 'planned', 'platform', 'proposed', 'raceway', 'steps', "motor"|
|highway|**not needed**|**not needed**|Remove feature if highway = ("motorway", "motorway_link", "footway", "bridleway") UNLESS bicycle = ("yes", "designated", "permissive", "destination") |Remove if highway= ("motorway", "motorway_link", "footway", "bridleway") UNLESS bicycle = ("yes", "designated", "permissive", "destination)|
|bicycle|"no"|"no"|"private", "no", "use_sidepath", "restricted"|"private", "no", "use_sidepath", "restricted"|
|service|"private"|"private"| "[str for str in values if "private" in str]", (in other words if 'private' is in a value for this tag, remove this feature) where values = network['service'].keys()|remove if key value contains substring "private"|




<br><br>
**Driving Filter** removed tags

| Tag | Pyrosm Values| OSMnx Values | osmextract|Proposed Filter|
|:----:|:----:|:----:|:---:|:---:|
|area| "yes" | "yes"| **None**|"yes"|
|access|**None**|"private"|"private", "no"|"private", "no"|
|highway| "cycleway", "footway", "path", "pedestrian", "steps", "track", "corridor", "elevator", "escalator", "proposed", "construction", "bridleway", "abandoned", "platform", "raceway"| "abandoned", "bridleway", "bus_guideway", "construction", "corridor", "cycleway", "elevator", "escalator", "footway", "path", "pedestrian", "planned", "platform", "proposed", "raceway", "service", "steps", "track"|'abandoned', 'bus_guideway', 'byway', 'construction', 'corridor', 'elevator', 'fixme', 'escalator', 'gallop', 'historic', 'no', 'planned', 'platform', 'proposed', 'cycleway', 'pedestrian', 'bridleway', 'path', 'footway', 'steps'|'abandoned', 'bus_guideway', 'byway', 'construction', 'corridor', 'elevator', 'fixme', 'escalator', 'gallop', 'historic', 'no', 'planned', 'platform', 'proposed', 'cycleway', 'pedestrian', 'bridleway', 'path', 'footway', 'steps', "track", "raceway", "service"|
|motor_vehicle| "no"|"no"|**None**|"no"|
|motorcar| "no"| "no"|**None**|"no"|
|service| "parking", "parking_aisle", "private", "emergency_access"| "alley", "driveway", "emergency_access", "parking", "parking_aisle", "private"| "[str for str in values if "private" in str]", (in other words if 'private' is in a value for this tag, remove this feature) where values = network['service'].keys()|"alley", "driveway", "emergency_access", "parking", "parking_aisle", "private"/[str for str in values if ”private” in str]|
<br><br>
**Driving & Service Filter** removed tags

| Tag | Pyrosm Values| OSMnx Values |osmextract|
|:----:|:----:|:----:|:---:|
|access|**None**|"yes"|*unspecified*|
|area|"yes"| "yes"|*unspecified*|
|highway|"cycleway", "footway", "path", "pedestrian", "steps", "track", "corridor", "elevator", "escalator", "proposed", "construction", "bridleway", "abandoned", "platform", "raceway"|"abandoned", "bridleway", "bus_guideway", "construction", "corridor", "cycleway", "elevator", "escalator", "footway", "path", "pedestrian", "planned", "platform", "proposed", "raceway", "steps", "track"|*unspecified*|
|motor_vehicle|"no"|"no"|*unspecified*|
|motorcar|"no"|"no"|*unspecified*|
|service|"parking", "parking_isle", "private", "emergency_access"|"emergency_access", "parking", "parking_aisle", "private"|*unspecified*|
|psv|"yes"|**None**|*unspecified*|
<br><br>
**Walking Filter** removed tags

| Tag | Pyrosm Values| OSMnx Values | osmextract|Proposed Filter|
|:----:|:----:|:----:|:---:|:---:|
|access|**None**|"private"|"private", "no"|"private","no"|
|area| "yes" | "yes"|**None**|"yes"|
|highway| "cycleway", "motor", "proposed", "construction", "abandoned", "platform", "raceway", "motorway", "motorway_link"|"abandoned", "bus_guideway", "construction", "cycleway", "motor", "planned", "platform", "proposed", "raceway"|'abandoned', 'bus_guideway', 'byway', 'construction', 'corridor', 'elevator', 'fixme', 'escalator', 'gallop', 'historic', 'no', 'planned', 'platform', 'proposed', 'raceway', 'motorway', 'motorway_link'|'abandoned', 'bus_guideway', 'byway', 'construction', 'corridor', 'elevator', 'fixme', 'escalator', 'gallop', 'historic', 'no', 'planned', 'platform', 'proposed', 'raceway', 'motorway', 'motorway_link', 'motor'|
|highway|**not needed**|**not needed**|Remove feature if highway = "cycleway" UNLESS feature also has foot = "yes", "permissive", "designated", "destination"|remove feature if highway=cycleway UNLESS feature has foot=("yes", "permissive", "designated", "destination")|
|foot| "no"| "no"|"private", "no", "use_sidepath", "restricted"|"no", "private", "use_sideapth", "restricted"|
|service|"private"| "private" | "[str for str in values if "private" in str]", (in other words if 'private' is in a value for this tag, remove this feature) where values = network['service'].keys()|	“[str for str in values if ”private” in str]”|

<br><br>.

## Creating Proposed Tag Filters

Here we show the code used to determine the final combination of proposed tag filters to use as our proposed filters within the tables above.
<br><br> 
This was done by creating a set of each packages tags filter and finding the setdifference and combinging this, for each network type. See below: 

### Drving Filter

```{python}
# Driving Filter (highway tag)
pyrosm_set_drive = set(["cycleway", "footway", "path", "pedestrian", "steps", "track", "corridor", "elevator", "escalator", "proposed", "construction", "bridleway", "abandoned", "platform", "raceway"])
OSMnx_set_drive = set(["abandoned", "bridleway", "bus_guideway", "construction", "corridor", "cycleway", "elevator", "escalator", "footway", "path", "pedestrian", "planned", "platform", "proposed", "raceway", "service", "steps", "track"])
osmextract_set_drive = set(['abandoned', 'bus_guideway', 'byway', 'construction', 'corridor', 'elevator', 'fixme', 'escalator', 'gallop', 'historic', 'no', 'planned', 'platform', 'proposed', 'cycleway', 'pedestrian', 'bridleway', 'path', 'footway', 'steps'])
proposed_set_drive = set(['abandoned', 'bus_guideway', 'byway', 'construction', 'corridor', 'elevator', 'fixme', 'escalator', 'gallop', 'historic', 'no', 'planned', 'platform', 'proposed', 'cycleway', 'pedestrian', 'bridleway', 'path', 'footway', 'steps', "track", "raceway", "service"]) 

# Things in osmextract filter, not in OSMnx/pyrosm filters
print(osmextract_set_drive.difference(pyrosm_set_drive))
# Things in osmextract filter, not in OSMnx/pyrosm filters
print(osmextract_set_drive.difference(OSMnx_set_drive))

# Things in OSMnx/pyrosm filters, not in osmextract
print(OSMnx_set_drive.difference(osmextract_set_drive))
# Things in OSMnx/pyrosm filters, not in osmextract
print(pyrosm_set_drive.difference(osmextract_set_drive))
```
<br><br>
The proposed filter is the **osmextract** filter with "track", "raceway", "service" keys added.
<br><br>

### Cycling Filter

```{python}
pyrosm_cycle = set(["footway", "steps", "corridor", "elevator", "escalator", "motor", "proposed", "construction", "abandoned", "platform", "raceway", "motorway", "motorway_link"])
OSMnx_cycle = set(["abandoned", "bus_guideway", "construction", "corridor", "elevator", "escalator", "footway", "motor", "planned", "platform", "proposed", "raceway", "steps"])
osmextract_cycle = set(['abandoned', 'bus_guideway', 'byway', 'construction', 'corridor', 'elevator', 'fixme', 'escalator', 'gallop', 'historic', 'no', 'planned', 'platform', 'proposed', 'raceway', 'steps'])
osmextract_UNLESS_cycle = set(["motorway", "motorway_link", "footway", "bridleway"])
osmextract_combined = set(['abandoned', 'bus_guideway', 'byway', 'construction', 'corridor', 'elevator', 'fixme', 'escalator', 'gallop', 'historic', 'no', 'planned', 'platform', 'proposed', 'raceway', 'steps', "motorway", "motorway_link", "footway", "bridleway"])

# Tags in osmextract filter not in OSMnx/pyrosm filters
print(osmextract_combined.difference(OSMnx_cycle))
# Tags in osmextract filter not in OSMnx/pyrosm filters
print(osmextract_combined.difference(pyrosm_cycle))

#Tags in OSMnx/pyrosm filters not in osmextract filters
print(OSMnx_cycle.difference(osmextract_combined))
#Tags in OSMnx/pyrosm filters not in osmextract filters
print(pyrosm_cycle.difference(osmextract_combined))
```

<br><br>
The proposed filter is the **osmextract** filter with "motor" key added.
<br><br>
Note this will still include the remove feature if highway=["motorway", "motorway_link", "footway", "bridleway"] UNLESS bicycle=["yes", "designated", "permissive", "destination"] clause.
<br><br>

### Walking Filter

```{python}
#Walking Filter
pyrosm_walk = set(["cycleway", "motor", "proposed", "construction", "abandoned", "platform", "raceway", "motorway", "motorway_link"])
OSMnx_walk = set(["abandoned", "bus_guideway", "construction", "cycleway", "motor", "planned", "platform", "proposed", "raceway"])
osmextract_walk = set(['abandoned', 'bus_guideway', 'byway', 'construction', 'corridor', 'elevator', 'fixme', 'escalator', 'gallop', 'historic', 'no', 'planned', 'platform', 'proposed', 'raceway', 'motorway', 'motorway_link'])
osmextract_UNLESS_walk = set("cycleway")
osmextract_combined = set(['abandoned', 'bus_guideway', 'byway', 'construction', 'corridor', 'elevator', 'fixme', 'escalator', 'gallop', 'historic', 'no', 'planned', 'platform', 'proposed', 'raceway', 'motorway', 'motorway_link', "cycleway"])

# Comparing OSMnx & Pyrosm filter
print(pyrosm_walk.difference(OSMnx_walk))
# Comparing OSMnx & Pyrosm filter
print(OSMnx_walk.difference(pyrosm_walk))

# Update pyrosm/OSMnx filters to match
pyrosm_walk.update(["bus_guideway", "planned"])
OSMnx_walk.update(["motorway_link", "motorway"])

# Comparing updated filters
print(pyrosm_walk.difference(OSMnx_walk))
# Comparing updated filters
print(OSMnx_walk.difference(pyrosm_walk))

# Comparing difference to osmextract filter
print(OSMnx_walk.difference(osmextract_walk))
# Comparing difference to osmextract filter
print(pyrosm_walk.difference(osmextract_walk))
```

<br><br>
We can see that the osmnextract filter is missing both "motor" and "cycleway".
<br><br>
For the proposed filter we will add "motor" but not cycleway, as this will be
taken care of by the proceeding highway=cycleway UNLESS foot=("yes", "designated", "permissive", "destination") clause.

<br><br>

### Proposed Data Aquisition Method. 

I am looking into how each package (osmextract, pyrosm, OSMnx) acquires OSM data so that we may use the most appropriate method moving forward.
<br><br>
Pyrosm collects pre-formated OSM extracts by custom user query from one of two providers, BBBike or Geofabrik.
<br><br>
It appears that Pyrosm takes all data for a user place query, before iteratively removing features from the parsed .osm.pbf file if they do not satisfy the filter conditions specified by the [filter documentation](https://github.com/HTenkanen/pyrosm/blob/1d93eb050a37e4564b56f49ef816e409b6ae658c/pyrosm/config/osm_filters.py#L1). Features that are removed from the file unless specified by the filter documentation, suggesting that some incorrect (*unknown*) features may still remain in the network if they are not caught by this filter. 
<br><br>
On the other hand, it seems that osmextract makes direct queries to the OSM API to retrieve network features that meet a specific SQL query. The network features returned from such a request will only be returned if they satisfy the SQL query passed to the OSM API, and as such only known features should be returned, unlike pyrosm. 
<br><br> 
From studying the OSM documentation, it seems this may take a similar approach to osmextract making direct queries to the OSM API but I need to study this further. 
<br><br>
I propose that we proceed utilizing the osmextract library, making use of the SQL query option so that we can specifically define queries (i.e. highway=cycleway UNLESS foot=yes, permissive, designated, destination etc...)
