---
title: "Transport regions in the UK"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Transport regions in the UK}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9,
  fig.height = 7.2
)
```

Within this vignette we show the Local Authority Districts (LADs) within the UK & Northern Ireland along with the transport regions for England

## Required library imports

```{r setup}
#library(openinfra)
library(sf)    # Simple Features (sf) library for storing geospatial data 
library(tmap)  # tmap library for interactive plots
library(dplyr) 
```
Note that these packages must already be installed before importing. To install these packages you can run the following lines: 
```
pkgs = c('sf', 'tmap', 'dplyr')
lapply(pkgs, library, character.only = TRUE)[length(pkgs)]
```
## Local authorities of the UK

Downloading, reading and plotting the Local Authority Districts (LADs) of the UK

```{r}
# Download data from the openinfra repo
url_regions_lad = "https://github.com/udsleeds/openinfra/raw/main/data-small/lads_joined_2021.geojson"

# Load data into a Simple Features frame
regions_lad = read_sf(url_regions_lad)

# Visualise the data
plot(regions_lad)
```
<br>
As can be seen the data contains a number of columns for each LAD including: 

* LAD21NM:      The name of each LAD
* Region_name:  The name of the region each LAD resides in
* Level:        The Authority level - either County, Combined authority or Unitary authority

<br>

### Interactive map of local authorities
Here we look to create an interactive plot of all local authority districts within UK & NI whereby clicking on each local authority brings up additional information associated with that local authority. <br> We can do this by changing the ```tmap_mode()``` from `'plot'` to `'view'`. Note that when in `tmap_mode('plot')` plots are not interactive
```{r}
# Set tmap mode to interactive
tmap::tmap_mode('view') # After this, all qtm() plots will be interactive
```
<br> Due to a [change](https://github.com/r-spatial/sf/issues/1649#issuecomment-860231045) in the spherical coordinates (S2) within the sf package that is known to cause issues ([1](https://github.com/r-tmap/tmap/issues/577), [2](https://github.com/r-tmap/tmap/issues/571)) when plotting geospatial data we make use of a known [fix](https://github.com/r-spatial/sf/issues/1649#issuecomment-860231045) to use the old coordinate system by setting `sf::sf_use_s2(use_s2 = FALSE)`
```{r}
# Here we create the  interactive plots
sf::sf_use_s2(use_s2 = FALSE)
regions_lad %>% tmap::qtm(title = 'LADs of UK & NI')
```
<br> Here we call another interactive map of all the LADs, but this time we have colour coded the LADs based on the region they reside in
```{r, fig.height=15.25, fig.width=9}
tmap_options(max.categories = 78) # We increase max categories so that every transport region has its own colour
regions_lad %>% tmap::qtm(fill = 'Region_name', title = 'LADs of UK & NI coloured by region')
```
<br><br>

## Transport regions within England

Downloading, reading and plotting the transport regions of England

```{r}
# Download data from the openinfra repo
url_transport_regions = "https://github.com/udsleeds/openinfra/raw/main/data-small/regions_new.geojson"

# Load data into a Simple Features frame
transport_regions = read_sf(url_transport_regions)

# Visualise transport regions of England
plot(transport_regions)
```

### Interactive map of transport regions within England
Here we look to create an interactive plot of all local transport regions within England. Again clicking on each transport region will highlight additional information, such as the specific LADs that reside within each transport region.
```{r}
transport_regions %>% tmap::qtm(title = 'Transport regions of England')
```

```{r, fig.height=15.25, fig.width=9}
# Here we create the interactive plot of England's transport regions
transport_regions %>% tmap::qtm(title = 'Coloured transport regions of England', fill = 'Region_name')
```


