---
title: "Transport regions in the UK"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Transport regions in the UK}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Info

Within this vignette we show the Local Authority Districts (LADs) within the UK & Northern Ireland along with the transport regions for England

## Required imports for analysis

```{r setup}
#library(openinfra)
library(sf)    # Simple features library for storing data 
library(tmap)  # tmap library for interactive plots
library(dplyr) 
```

## Local authorities of the UK

Downloading, reading and plotting the local authority districts of the UK

```{r}
# Download data from the openinfra repo
url_regions_lad = "https://github.com/udsleeds/openinfra/raw/main/data-small/lads_joined_2021.geojson"
# Load data into a Simple Features frame
regions_lad = read_sf(url_regions_lad)
# Visualise the local authority districts of the UK
plot(regions_lad)
```

```{r}
# Set our tmap package mode to _interactive_ to create interactive plots
tmap::tmap_mode('view') # After this call all qtm plots will be interactive 
```

```{r}
# Here we create the  interactive plots
sf::sf_use_s2(use_s2 = FALSE)
regions_lad %>% select('geometry') %>% tmap::qtm()
```

<br>[Interactive](https://github.com/udsleeds/openinfra/blob/main/data-small/lads_joined_2021.geojson) plot of the local authorities within the UK <br>

## Transport regions within England

Downloading, reading and plotting the transport regions of England

```{r}
# Download data from the openinfra repo
url_transport_regions = "https://github.com/udsleeds/openinfra/raw/main/data-small/regions_new.geojson"
# Load data into a Simple Features frame
transport_regions = read_sf(url_transport_regions)
# Visualise transport regions of England
plot(transport_regions)
```

```{r}
# Here we create the interactive plot of Englands transport regions
transport_regions %>% select(geometry) %>% tmap::qtm()
```


<br>[Interactive](https://github.com/udsleeds/openinfra/blob/main/data-small/regions_new.geojson) plot of transport regions within England
