---
title: "SOTM 2022 Workshop - OSM for sustainable transport planning: getting started"
author: "James Hulse"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SOTM2022 Workshop - OSM for sustainable transport planning: getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The [workshop](https://2022.stateofthemap.org/sessions/JEXYHK/) is aimed at everyone interested in using OpenStreetMap (OSM) to support sustainable transport planning, in professional or advocacy contexts through recategorisation of OpenStreetMap (OSM) transport infrastructure data. The focus will be on getting started, identifying, visualizing, and analyzing key tags and identifying gaps in walking, cycling, and wheeling networks. Participants will place OSM data in the context of other data sources to identify its unique advantages.

The workshop is designed to be practical and covers the following question:

1.  What gaps in transport evidence can OSM data fill in different countries?
2.  Which elements and tags are most important for transport planning (and how to get the data in reproducible data science environments)?
3.  How to filter based on different tags and their values?
4.  How to use OSM datasets to visualize transport systems in static and interactive maps?
5.  How to add value to OSM data and communicate results using R, Python and A/B Street, and osm2streets?

In order to take the most out of the workshop, we recommend installing R, RStudio, and the required packages before the session. To do this, follow "Installation" section below. Moreover, to make sure that everything is installed correctly and you're ready to go -- run a chunk of code in the "Installation Check" section! Check [this issue on GitHub](https://github.com/udsleeds/openinfra/issues/98) (you will need an account to make use of it) to find out what you should get after running that code or, alternatively, let us know if you have not :)

Finally, this workshop does not assume any prior knowledge of R or programming. Yet, it would be useful to familiarise yourself with it prior to the workshop. A good starting point is [RStudio's Beginners guide](https://education.rstudio.com/learn/beginner/) and its [Gentle Introduction to Tidy Statistics in R](https://www.rstudio.com/resources/webinars/a-gentle-introduction-to-tidy-statistics-in-r/).

```{=html}
<!-- This workshop covers: 
* How to download and get OSM data from the R package `osmextract`
* Basics on how the `osmextract` package works
* How to visualise different types of network infrastructure
* How to add value to OSM infrastructure data through the `openinfra` package -->
```
# Preparation

## Installing R and RStudio

This workshop will be carried out using R. If you are new to R, you should install base R and RStudio before the workshop begins. <br> There is a great [R and RStudio installation guide](https://rstudio-education.github.io/hopr/starting.html) (Section A.1) by Garrett Grolemund, within [Hands-On Programming with R](https://rstudio-education.github.io/hopr/) for Windows, Mac, and Linux devices.

## Installing Packages

A number of packages will be used within this workshop, though namely [`osmextract`](https://github.com/ropensci/osmextract) and [openinfra](https://github.com/udsleeds/openinfra).

Here are some of the additional packages we will make use of within this workshop:

```{r install_dep, eval = F}
pkgs = c("tmap",
         "sf",
         "dplyr",
         "remotes"
         )
```

These can be installed with:

```{r install_okgs, message=FALSE, eval = F}
install.packages(pkgs, repos = "http://cran.us.r-project.org")
```

You can install the development versions of the `osmextract` and `openinfra` packages as follows:

```{r install_dev, message=FALSE, eval = F}
# Enter the lines below into your console to install!
remotes::install_github("ropensci/osmextract", force = T)
remotes::install_github("udsleeds/openinfra")
```

Each package can then be loaded iteratively with `library(tmap)`, `library(sf)`, etc, or all at once with:

```{r load_pkgs, message=FALSE, eval = F}
pkgs = c(pkgs, "osmextract", "openinfra")
lapply(pkgs, library, character.only = TRUE)[length(pkgs)]
```

## Installation Check

If installation has occurred correctly, running the code below locally (on your own device) should successfully create two plots: <br> One 5km radius circular buffer of the infrastructure network around Leeds (UK) City Centre, and <br> Secondly, a test plot just outside of the Institute of Transport Studies (ITS), at the University of Leeds.

```{r install_check, message=FALSE, dpi=300, out.width = '100%', eval = F}
# Pre-installed package data for central Leeds
data = openinfra::sotm_data
# Test osmextract file (ITS Leeds = Institute of Transport Studies, University of Leeds)
test_data = osmextract::oe_get(place = "ITS Leeds")

# Plot central Leeds
tmap::tm_shape(data |> dplyr::select("highway")) + 
  tmap::tm_lines(col = "highway", title.col = "OSM Highways") + 
  tmap::tm_layout( legend.bg.alpha = 0.5, legend.bg.color = "white")

# Plot ITS
tmap::tm_shape(test_data |> dplyr::select("highway")) + 
  tmap::tm_lines(col = "highway", title.col = "OSM Highways") + 
  tmap::tm_layout( legend.bg.alpha = 0.5, legend.bg.color = "white")
```

## Why these packages?

The [`osmextract`](https://github.com/ropensci/osmextract) package is an open-source project, which allows for fast downloading and importing of OSM infrastructure data from one of three OSM providers. We can view all `osmextract` available providers with the following:

```{r show_providers, message=FALSE, eval = F}
osmextract::oe_providers() 
```

<br> `osmextract` uses [Geofabrik](http://download.geofabrik.de/), [BBBike](https://download.bbbike.org/osm/), and [openstreetmap_fr](https://download.openstreetmap.fr/extracts/) as OSM data providers. These providers extract, select, and process geospatial data from OSM in .osm.pbf file format. The .pbf file format results in much faster read/write times and reduces filesizes, more information can be found on the [OSM wiki](https://wiki.openstreetmap.org/wiki/PBF_Format). <br><br> The [openinfra](https://github.com/udsleeds/openinfra) package, another open-source project, contains several functions that recategorises OSM infrastructure data from the default key=tag values to friendlier visuals.

[`tmap`](https://r-tmap.github.io/tmap/index.html) is a fantastic package for visualising geospatial data either statically or interatively.

[`sf` package](https://r-spatial.github.io/sf/) supports the representation of spatial vector data as simple features which, consequentially, makes spatial analysis much more accessible. This package is, actually, a backbone to everything we currently do :)

# Getting data

In this section we will use `osmextract` to download OSM data that will be used throughout this workshop.

The `osmextract` package comprises the following core functions (also see a [vignette introducing `osmextract`](https://docs.ropensci.org/osmextract/articles/osmextract.html?q=file%20path#main-package-functions)): 

| Function | Description |
|:------:| -------- |
|`oe_providers()`|Shows a data frame of currently supported OSM data providers|
|`oe_match()`|  Matches the input place query with a url from one of the OSM providers|
| `oe_download()`| Downloads data from OSM providers|
| `oe_vectortranslate()` |  Converts between .pbf (providers default) and .gpkg file formats |
| `oe_read()` | Reads downloaded .pbf and .gpk files into R|
| `oe_get()` |  Performs all of the above in a single function|

To start off we will use a generic `oe_get()` function to download data for Leeds, UK. It might take a while. If you struggle downloading the data or it takes long to compute, then you might use the example dataset (e.g. `data`) for central Leeds. It is a much smaller dataset, thus making computing much quicker. 

```{r, eval = F}
# A useful function is `oe_match_pattern()` to search for patterns in a provider's database. It is helpful in indicating a correct region name for a given provider.
osmextract::oe_match_pattern("leeds") 

leeds = osmextract::oe_get(
  place = "Leeds", # Note capital 'L'!
  provider = "bbbike", # Indicates the provider; default is geofabrik
  layer = "lines", # Default; returns linestring geometries (highways, waterways, aerialways) 
  force_download = TRUE, # Updates the previously downloaded .osm.pbf file (default is FALSE)
  force_vectortranslate = TRUE # Forces the vectorization of a .pbf file to .gpbf even if there is a .gpbf file with the same name (default = FALSE)
)

# have a look at the data
leeds |> dplyr::glimpse()

```

## Tags

In the previous example we've downloaded OSM data but it does not necessarily contain all the information we need. Indeed, it only has 10 columns and only `highway` column is directly useful for active travel infrastructure planning!

In order to get additional tags we need to specify this as an additional argument before running `oe_get()`.

```{r, eval = F}
# so let's define the extra tags we want

et = c("kerb", "width", "sidewalk", "cycleway", "footway", "lit", "wheelchair", "maxspeed", "foot", "access", "service", "bicycle", "oneway")

leeds = osmextract::oe_get(
  place = "Leeds", # Note capital 'L'!
  provider = "bbbike", # Indicates the provider; default is geofabrik
  layer = "lines", # Default; returns linestring geometries (highways, waterways, aerialways) 
  force_download = TRUE, # Updates the previously downloaded .osm.pbf file (default is FALSE)
  force_vectortranslate = TRUE, # Forces the vectorization of a .pbf file to .gpbf even if there is a .gpbf file with the same name (default = FALSE)
  extra_tags = et # non-default tags to download
)

# have a look at the data
leeds |> dplyr::glimpse()

# also it's useful to have a look at the downloaded data by plotting it 
# it helps to 
tmap::tm_shape(leeds |> dplyr::filter(!is.na("highway")))+
  tmap::tm_lines(col = "highway")
```

Strangely (or not!) our Leeds data actually represents the whole of West Yorkshire and not only Leeds as we requested. Such bugs happen, that's why it's important to inspect data before proceeding further.

## Spatial subsetting

Next we will subset Leeds city from `leeds` dataset. Spatial subsetting is one of the essential operations in geocomputation, so if you want to learn more about this, check out ["Spaptial subsetting"](https://geocompr.robinlovelace.net/spatial-operations.html#spatial-subsetting) section in the Geocomputation with R (Lovelace et al., 2022).

We will use a combination of `mapview` and `mapedit` packages to manually define an area, get its coordinates to perform spatial subsetting.

```{r, eval = F}
map_leeds = mapview::mapview(leeds) # create an object containing interactive map  
map_leeds_edit = mapedit::editMap(map_leeds) # use `mapedit` package to draw a new polygon
box = sf::st_bbox(map_leeds_edit$drawn$geometry) # extract the coordinates of a new polygon
box
```

Now we have coordinates of an area we are interested in. From here on there are two ways for getting OSM data for that. One is to use `oe_get()` again and give it an additional `boundary` and `boundary_type` arguments.
```{r, eval = F}
leeds_defined = osmextract::oe_get(
  place = "Leeds", # Note capital 'L'!
  provider = "bbbike", # Indicates the provider; default is geofabrik
  layer = "lines", # Default; returns linestring geometries (highways, waterways, aerialways) 
  force_download = TRUE, # Updates the previously downloaded .osm.pbf file (default is FALSE)
  force_vectortranslate = TRUE, # Forces the vectorization of a .pbf file to .gpbf even if there is a .gpbf file with the same name (default = FALSE)
  extra_tags = et, # non-default tags to download
  boundary = box,
  boundary_type = "clipsrc"
)
```

Alternatively, we can subset the `leeds` data, so we do not have to download data again. 
```{r, eval = F}
leeds_defined2 = leeds[map_leeds_edit$drawn$geometry, 
                       op= sf::st_intersects] 
```

**Task**: 

1. Decide which tags you would like to download (be selective!)
2. Download OSM data for a place of your choice by modifying the code above. 
3. Bonus: subset your dataset and plot it. 

Note: not every city/town/region might be directly queried using osmextract.

# Data exploration

As alluded in the previous section, it is important to understand the data before using it for further (spatial) analysis or inform decision-making.

In this section you will be introduced to data filtering based on OSM tags and how to visualise them.

## Data filtering

Our data contains a lot of information, but not all of it might be needed. For example, active travel includes cycling, walking, and wheeling, but you may only be interested in cycling. In this case it would be reasonable to contain only data that is related to cycling. In other words, we may not need data on waterways or motorways as one cannot cycle on them.

```{r, eval = F}
# filter out all the rows that do not contain highways
leeds_highways = leeds_defined2 |> 
  dplyr::filter(!is.na(highway))
```

Our data now only contains highways in Leeds, but we want to create a new dataset containing highway=cycleway only and then visualise them by plotting on a map.
```{r, eval = F}
# filter and select
leeds_cycleway = leeds_highways |> 
  dplyr::filter(highway == "cycleway") |> # filtering based on a key value
  dplyr::select(highway)

# plot
leeds_cycleway |> plot()
```

Of course, highway=cycleway does not represent an entire cycling infrastructure in the given region. For this we need to apply more complex (conditional) filtering to our data.

For example, you may know that it is illegal to cycle on certain roads, such as motorways in the UK. Also, it may be a case that a cycleway is not mapped separately but, rather, has been tagged as part of another highway, such as tertiary road, using `bicycle=*` tag.

```{r, eval = F}
# To filter data based on these requirements we will:

leeds_cycle = leeds_highways |> 
  dplyr::filter(highway == "cycleway" |
                bicycle %in% c('yes'))  

# let's plot our new cycle network
leeds_cycle["geometry"] |> plot()
```

**Task**:

1. Think what other tags and/or their values might be useful to define cycling network. Use them to further refine the network and plot it.
2. Do you think your new network accurately represents the existing network?
3. What do you think are potential challenges in defining a network? (eg. missing data)
4. Bonus: try to define a pedestrian network.

## Pre-defined networks

It's not easy to define a network and, once conceptualised, a lack of data might still lead to innacurate representation, thus limiting further analysis and application.

Nevertheless, it's still useful to have a network defined. For instance, `osmextract` does offer an opportunity to download pre-defined walking, cycling, or driving networks. Yet, if you choose to do this, do check the code used to create the network to ensure that you understand how the network has been defined and that it works for your needs.

```{r, eval = F}
# to download cycling network as defined by `osmextract`
osme_cycle = osmextract::oe_get_network(place = "Leeds",
                                        mode = "cycling")
```

```{r, eval = F}
# plot cycling network
osme_cycle["geometry"] |>
  plot()
```

# Visualisation

In this workshop we've been visualizing our data to improve our understanding. Up to this point we've been visualising data statically using base R `plot()` function. However, there is a powerful `tmap` library that can be used to create static and interactive plots to communicate your findings.

```{r, eval = F}
# equivalent to the earlier plotted map
# Note: it might be useful to store your plot as a tmap object that can be called at any point.

osme_cycle_plot = tmap::tm_shape(osme_cycle["geometry"])+
  tmap::tm_lines()
```

To plot an interactive map you have to specify this using `tmap_mode()` function. 

```{r, eval = F}
tmap::tmap_mode("view") # now all the plots will be interactive
tmap::tm_shape(osme_cycle)+
  tmap::tm_lines("highway")
```

**Task**:

1. Plot an interactive map of a network you defined in the previous task.
2. Create a `tmap` object of a static map. Tip: run `?tmap_mode()` to access function's documentation.
3. Bonus: Save your new tmap object. Tip: explore `tmap_save()` function.

# OpenInfra

In this section we are going to explore OpenInfra package and some of the functions it currently contains.

One of the functions that the pakcage has is `oi_recode_road_class()`. It recategorises OSM highways based on the Chan and Cooper's (2019) work in which they propose that road class can be used to infer motor traffic volume. Road traffic volume is an important factor in discouraging cycling. 

```{r}
leeds_high_rec = leeds_defined2 |> 
  openinfra::oi_recode_road_class()
```

```{r}
tmap::tm_shape(leeds_high_rec)+
  tmap::tm_lines("oi_road_desc")
```

























