---
title: "census+pct+osm+leeds"
output: md_notebook
---

```{r}
# load up the libraries
library(readr)
library(osmextract) 
library(tidyverse)
library(sf)
```

```{r}
wy_walking <- readRDS("/Users/gretatimaite/Desktop/openinfra/test-code/wy_walking.Rds")
leeds_pct <- readRDS("/Users/gretatimaite/Desktop/openinfra/test-code/leeds_pct.Rds")
leeds_joined_lsoa <- readRDS("/Users/gretatimaite/Desktop/openinfra/test-code/leeds_joined_lsoa.Rds")
```


## linking pct data with Leeds Census 2011
```{r, message = FALSE}
# working with lsoa data
# travel_foot_leeds_lsoa <- read_csv("~/Desktop/Method of travel to work On foot % 2011 lsoa.csv")
travel_foot_leeds_lsoa

# saving as .Rds
# saveRDS(travel_foot_leeds_lsoa,"travel_foot_leeds_lsoa.Rds")

# checking if leeds_pct$geo_code matches travel_foot_leeds_lsoa$NAME
leeds_pct$geo_code %in% travel_foot_leeds_lsoa$NAME %>% table() # TRUE 482

```

### pct + traveling on foot (%)
```{r}
# Cleaning data -- extracting digits from the `Method of travel to work: On foot %|2011` column

travel_foot_leeds_lsoa1 <- travel_foot_leeds_lsoa %>%
  mutate(foot_census = readr::parse_number(travel_foot_leeds_lsoa$"Method of travel to work: On foot %|2011")) # using `parse_number` to extract digits, which will be added to a new column called "value"

# there were 3 parsing failures. Let's check which ones failed
travel_foot_leeds_lsoa1 %>% filter(is.na(foot_census)) # not sure why they failed as I don't see any differences compared to other rows
# I'll fix this manually for now but I'll need to come up with more nuanced code to avoid parsing issues...
  
travel_foot_leeds_lsoa1$foot_census[travel_foot_leeds_lsoa1$NAME == "E01011375"] <- 7.2
travel_foot_leeds_lsoa1$foot_census[travel_foot_leeds_lsoa1$NAME == "E01011564"] <- 1.7
travel_foot_leeds_lsoa1$foot_census[travel_foot_leeds_lsoa1$NAME == "E01011670"] <- 12.7

# double-checking if there are any NAs left
travel_foot_leeds_lsoa1 %>% filter(is.na(value)) # none
```

### joining pct + census

```{r, message = FALSE}
# now we can join `leeds_pct` and `travel_foot_leeds_lsoa1` dataframes

leeds_joined_lsoa <- left_join(leeds_pct,
                          travel_foot_leeds_lsoa1,
                          by = c(geo_code = "NAME"))
# checking if we have `value` column
leeds_joined_lsoa %>% select(foot_census) # good :)
```

### plotting leeds_joined with osm
```{r}
leeds_joined_lsoa %>% select(foot_census) %>% plot

st_geometry(leeds_pct) %>% plot(reset = FALSE)
leeds_joined_lsoa %>% select(foot_census) %>% plot(add = TRUE)
wy_walking %>% filter(highway == "footway") %>% plot(add = TRUE,
                                                     col = "white")

# sf_col_table(leeds_joined_lsoa, "foot_census")

```

### pct + economically active (%)
Source: https://observatory.leeds.gov.uk/economy-and-employment/map/

What 'economically active' means?
Economically active ->	Employed ->	Includes part-time work
Economically active ->	Unemployed ->	Without a job - available and seeking work
Economically inactive ->	Economically inactive ->	Not seeking work - includes sick, retired, full-time students, looking after family

Question: 

- how is "Economically active: In employment" measured? If it says that '60%' does it refer to the total population of an area or to its sample (i.e., the one which is economically active)? 

```{r, message = FALSE}
# working with lsoa data
# econ_active_leeds_lsoa <- read_csv("~/Desktop/Economically active % 2011 lsoa.csv")
econ_active_leeds_lsoa

# save as .Rds
# saveRDS(econ_active_leeds_lsoa,
#         "econ_active_leeds_lsoa.Rds")

# checking if leeds_pct$geo_code matches travel_foot_leeds_lsoa$NAME
leeds_pct$geo_code %in% travel_foot_leeds_lsoa$NAME %>% table() # TRUE 482

```

```{r}
# Cleaning data -- extracting digits from the `Economically active %|2011` column

econ_active_leeds_lsoa1 <- econ_active_leeds_lsoa %>%
  mutate(econ_active_census = readr::parse_number(econ_active_leeds_lsoa$"Economically active %|2011")) # using `parse_number` to extract digits, which will be added to a new column called "value"

# there were 3 parsing failures. Let's check which ones failed
econ_active_leeds_lsoa1 %>% filter(is.na(econ_active_census)) # It's interesting that the same rows failed to parse as in foot_census case. 
  
econ_active_leeds_lsoa1$econ_active_census[econ_active_leeds_lsoa1$NAME == "E01011375"] <- 63.7
econ_active_leeds_lsoa1$econ_active_census[econ_active_leeds_lsoa1$NAME == "E01011564"] <- 75.2
econ_active_leeds_lsoa1$econ_active_census[econ_active_leeds_lsoa1$NAME == "E01011670"] <- 50.6

# double-checking if there are any NAs left
econ_active_leeds_lsoa1 %>% filter(is.na(value)) # none
```

### joining leeds_joined + census on economically active %

```{r, message = FALSE}
# now we can join `leeds_joined` and `econ_active_leeds_lsoa1`

leeds_joined_lsoa <- left_join(leeds_joined_lsoa,
                          econ_active_leeds_lsoa1,
                          by = c(geo_code = "NAME"))

# saving data
# saveRDS(leeds_joined_lsoa,
#         "leeds_joined_lsoa.Rds")

# checking if we have `value` column
# leeds_joined_lsoa %>% select(econ_active_census) 

```

### plotting

```{r}
leeds_joined_lsoa %>% select(econ_active_census) %>% plot

st_geometry(leeds_pct) %>% plot(reset = FALSE)
leeds_joined_lsoa %>% select(econ_active_census) %>% plot(add = TRUE)
wy_walking %>% filter(highway == "footway") %>% plot(add = TRUE,
                                                     col = "white")
```
