
```{r, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```


# Get website working

Currently we have a basic website at https://udsleeds.github.io/openinfra/

Aim: make it better, add placeholder articles, make it autobuild.

Starting point: get website built locally.

```{r}
pkgdown::build_site()
pkgdown::preview_site()
```

The previous version of the homepage can be found at:

```{r}
file.edit("test-code/README-website.Rmd")
```

Test package locally:

```{r}
devtools::check()
```

Make the package ignore files:

```{r}
usethis::use_build_ignore(files = "GISRUK_paper")
usethis::use_build_ignore(files = "test-code")
usethis::use_build_ignore(files = c(
  "gm.Rds",
  "mers.Rds",
  "osm_highway_cycleway_left.Rds",
  "wy.Rds",
  "README.Rmd",
  "README_files",
  "code",
  "data-small",
  "docs",
  "figures",
  "mcr.xcf",
  "wy_pct.Rds",
  "wy_walking.Rds"
  )
  )
```



Set-up webite on GitHub actions...

```{r}
?usethis::use_github_action_check_standard
?pkgdown::build_site_github_pages
usethis::use_github_action_check_standard()
usethis::use_tidy_description()
```

When we have the standard actions working we'll set-up pkgdown with:

```{r}
use_github_action("pkgdown")
```

Info on pkgs set-up: https://r-pkgs.org/

Create an article called Get Started:

```{r}
usethis::use_vignette(name = "openinfra", title = "Getting started with open data on transport infrastructure")
```






