---
title: "West Yorkshire OSM"
---

useful git commands:

git

git status

git checkout -b gretatest

git commit -am "test change"

git status

git show

git push --

git pull 

git branch -v

git log

```{r}
library(osmextract) 
library(tidyverse)
library(sf)
```

```{r}
# Aim: get cycleways in west yorkshire

# oe_match_pattern("Yorkshire")
# region_name <- "West Yorkshire"
# wy <- osmextract::oe_get(place = region_name,
#                          layer = "lines",
#                          force_download = TRUE,
#                          force_vectortranslate = TRUE) # importing West Yorkshire data
# 
# wy_cycling_net <- osmextract::oe_get_network(place = region_name,
#                                             mode = "cycling"
#                                              ) # importing WY cycling network; definition needs further exploration

# saveRDS(wy, 
#         "wy.Rds")
# saveRDS(wy_cycling_net,
#         "wy_cycling_net.Rds")

wy %>% glimpse()
wy_cycling_net %>% glimpse()

# wy %>% pull(z_order) %>% unique() # checking what it means; has something to do with storing in the SQL database. Will delete for now. Link: https://wiki.openstreetmap.org/wiki/Layer
wy_short <- wy %>% select(osm_id, 
                          name,
                          highway,
                          other_tags)
nrow(wy) == nrow(wy_short)

# wy_cycling_net$access %>% unique # might be useful to figure out which roads are actually useful for cyclists? i.e., filtered permeability? https://wiki.openstreetmap.org/wiki/Key:access
# wy_cycling_net$service %>% unique # https://wiki.openstreetmap.org/wiki/Key:service
wy_cn_short <- wy_cycling_net %>% 
  select(osm_id,
         name,
         highway,
         access,
         bicycle,
         service,
         other_tags
         )
nrow(wy_cn_short) == nrow(wy_cycling_net)

# wy_short %>% select(highway) %>% plot(main = "wy_short")
# wy_cn_short %>% select(highway) %>% plot(main = "wy_cn_short")
```

## **Andrea's comment**:
I checked the script that you mentioned (i.e. cycleways.R) and I have
one comment. The object osm_highway (line 8) includes all "OSM ways"
in the IoW and I think that the ratio nrow(osm_highway) /
nrow(osm_highways_cycling) does not represent a fair comparison (since
you are comparing ways and roads). If you want to estimate the ratio
of (cycleable streets) / (all streets in Iow), you can run:

osm_highway = osmextract::oe_get(place = "Isle of Wight", query =
"SELECT * FROM lines WHERE highway IS NOT NULL", quiet = TRUE)
osm_highways_cycling = osmextract::oe_get_network(place = "Isle of
Wight", mode = "cycling")

and you will notice that

nrow(osm_highway) / nrow(osm_highways_cycling) = 1

since, according to usual OSM tags, bicycle transportation is allowed
everywhere but for motorways, which are missing in the IoW. If you
want to create a more restrictive definition for a cycling network
(i.e. exclude primary or secondary roads), then you can modify the
code behind oe_get_network (see also the old PR:
https://github.com/ropensci/osmextract/pull/224).


```{r}
# applying Andrea's comment to West Yorkshare 

# osm_highway = osmextract::oe_get(place = region_name, query =
# "SELECT * FROM lines WHERE highway IS NOT NULL", quiet = TRUE, force_download = TRUE,
#                          force_vectortranslate = TRUE)
# osm_highways_cycling = osmextract::oe_get_network(place = region_name, mode = "cycling", force_download = TRUE,
#                          force_vectortranslate = TRUE)
nrow(osm_highway)
#> 180661
nrow(osm_highways_cycling)
#> 167853
c(nrow(osm_highway) == nrow(osm_highways_cycling), # FALSE
  nrow(osm_highway) / nrow(osm_highways_cycling) ) # 1.076305 ratio
# > 0.000000 1.076305

# checking if NULLs are NAs 
wy_cn_short %>% nrow()
#> 167853 
wy_cn_short %>% filter(!is.na(highway)) %>% nrow() # double checking if there are non NAs in the `highway` column
#> 167853 
wy_short %>% select(highway) %>% nrow() # includes ALL ways (inlcuding NA)
#> 233598
wy_short %>% filter(!is.na(highway)) %>% nrow() 
#> 180661

wy_short %>% filter(!is.na(highway)) %>% nrow() == nrow(osm_highway) # so NAs are NULLs as expected
# > TRUE

# Alternative way to figure out NAs
# wy_short %>% pull(highway) %>% is.na() %>% sum() # we have 52937 NAs
# wy_short %>% pull(highway) %>% table %>% sum() # 180661 NAs not included in the table
# ( (wy_short %>% select(highway) %>% nrow()) - (wy_short %>% pull(highway) %>% table %>% sum()) ) == wy_short %>% pull(highway) %>% is.na() %>% sum() # double checking

wy_short %>% filter(!is.na(highway)) %>% nrow() == wy_cn_short %>% nrow()  # we still have unequal number of rows; I think it might have something to do with highway cycling network definition
# > FALSE
(wy_short %>% filter(!is.na(highway)) %>% nrow()) - (wy_cn_short %>% nrow()) # we have 12808 more ways in wy_short compared to wy_cn_short
# > 12808

wy_short %>% pull(highway) %>% table()
wy_cn_short %>% pull(highway) %>% table()
# .
#      bridleway   bus_guideway         busway   construction       corridor       cycleway        footway  living_street       motorway  motorway_link             no           path     pedestrian 
#           1268             14              7            339             48           3250          36956            392            577            393              6           7286            801 
#        primary   primary_link       proposed        raceway    residential           road      secondary secondary_link        service       services          steps       tertiary  tertiary_link 
#           3415            321             78             17          44638             27           1470             43          53252              2           2685           5174            143 
#          track          trunk     trunk_link   unclassified 
#           7958           2888            338           6875 
# .
#      bridleway         busway       cycleway        footway  living_street       motorway  motorway_link           path     pedestrian        primary   primary_link    residential           road 
#           1260              6           3233          35543            368            571            389           6928            783           3415            321          44379             27 
#      secondary secondary_link        service       services       tertiary  tertiary_link          track          trunk     trunk_link   unclassified 
#           1470             43          46584              2           5156            142           7205           2878            338           6812 

# so I'll filter data based on the highway tags present in cycling network dataset `wy_cn_short`
highway_tags <- c("bridleway", "busway", "cycleway", "footway", "living_street", "motorway", "motorway_link", "path", "pedestrian", "primary", "primary_link", "residential", "road", "secondary", "secondary_link", "service", "services", "tertiary", "tertiary_link", "track", "trunk", "trunk_link",  "unclassified")
wy_checking <- wy_short %>% filter(highway %in% highway_tags)
wy_checking %>% nrow() 
# > 177474
wy_checking %>% nrow() == wy_cn_short %>% nrow() 
# > FALSE
( wy_checking %>% nrow() ) - ( wy_cn_short %>% nrow() ) # still significantly more lines are present in the wy_short dataset; cycling network must have had some more refined filtering; 
# > 9621



# wy_checking %>% select(highway) %>% plot(main = "wy_checking")
# wy_cn_short %>% select(highway) %>% plot(main = "wy_cn_checking")
```

It's interesting that Andrea's comment is applicable to IoW but not WY, I reckon it depends on the types of ways/roads present in the area and how well they are tagged. 

## tags

```{r}

? osmextract::oe_get_keys(wy_short)
all_tags <- osmextract::oe_get_keys(wy_short) 

# maxspeed
all_tags %>% str_detect("maxspeed") 
speed = all_tags %>% str_detect("maxspeed")
wy_short[speed, ] %>% select(highway) %>% plot()

# speed limits
speed_data <- wy_short %>% mutate(speed = case_when(
  str_detect(other_tags,"10 mph") ~ 10,
  str_detect(other_tags,"20 mph") ~ 20,
  str_detect(other_tags,"30 mph") ~ 30,
  str_detect(other_tags,"40 mph") ~ 40,
  str_detect(other_tags,"50 mph") ~ 50,
  str_detect(other_tags,"60 mph") ~ 60,
  str_detect(other_tags,"70 mph") ~ 70,
  str_detect(other_tags,"80 mph") ~ 80,
  str_detect(other_tags,"90 mph") ~ 90,
  str_detect(other_tags,"100 mph") ~ 100)
                                           ) %>% 
  filter(!is.na(speed)) # getting rid of NAs

# how many rows
speed_data %>% nrow()
# > 19754
speed_data %>% pull(speed) %>% unique # double checking if all limits are present
speed_data %>% select(speed) %>% plot()

# leeds subset

# wy_pct = pct::get_pct_zones(region = "west-yorkshire")
# sf::sf_use_s2(FALSE)
# saveRDS(wy_pct,
#         "wy_pct.Rds")
leeds <- wy_pct %>% filter(lad_name == "Leeds")

# plotting speed data on leeds
st_geometry(leeds) %>% plot(reset = FALSE, 
                            col = "gray",
                            main = "speed limits in Leeds (gray area)")
speed_data['speed'] %>% plot(add = TRUE, 
                             legend = TRUE
                             )



```


## things to explore further:
- definition of cycling network in osm package
  - does it allign with LTN 1/20 
- how 'access' column can be used to evaluate the quality of the cycling network?

















