---
title: "West Yorkshire OSM"
---

useful git commands:

git

git status

git checkout -b gretatest

git commit -am "test change"

git status

git show

git push --

git pull 

git branch -v

git log

```{r}
library(osmextract) 
library(tidyverse)
library(sf)
```

```{r}
# Aim: get cycleways in west yorkshire

# oe_match_pattern("Yorkshire")
# region_name <- "West Yorkshire"
# wy <- osmextract::oe_get(place = region_name,
#                          layer = "lines",
#                          force_download = TRUE,
#                          force_vectortranslate = TRUE) # importing West Yorkshire data
# 
# wy_cycling_net <- osmextract::oe_get_network(place = region_name,
#                                             mode = "cycling"
#                                              ) # importing WY cycling network; definition needs further exploration

# saveRDS(wy,
#         "wy.Rds")
# saveRDS(wy_cycling_net,
#         "wy_cycling_net.Rds")

wy <- readRDS("wy.Rds")
wy_short <- readRDS("wy_short.Rds")
wy_cycling_net <- readRDS("wy_cycling_net.Rds")
wy_cn_short <- readRDS("wy_cn_short.Rds")
wy_pct <- readRDS("wy_pct.Rds")
leeds <- wy_pct %>% filter(lad_name == "Leeds")
wy_walking <- readRDS("wy_walking.Rds")


wy %>% glimpse()
wy_cycling_net %>% glimpse()

# wy %>% pull(z_order) %>% unique() # checking what it means; has something to do with storing in the SQL database. Will delete for now. Link: https://wiki.openstreetmap.org/wiki/Layer
wy_short <- wy %>% select(osm_id, 
                          name,
                          highway,
                          other_tags)
nrow(wy) == nrow(wy_short)

# saveRDS(wy_short,
#         "wy_short.Rds")

# wy_cycling_net$access %>% unique # might be useful to figure out which roads are actually useful for cyclists? i.e., filtered permeability? https://wiki.openstreetmap.org/wiki/Key:access
# wy_cycling_net$service %>% unique # https://wiki.openstreetmap.org/wiki/Key:service
wy_cn_short <- wy_cycling_net %>% 
  select(osm_id,
         name,
         highway,
         access,
         bicycle,
         service,
         other_tags
         )
nrow(wy_cn_short) == nrow(wy_cycling_net)

# wy_cn_short %>% saveRDS("wy_cn_short.Rds")

# wy_short %>% select(highway) %>% plot(main = "wy_short")
# wy_cn_short %>% select(highway) %>% plot(main = "wy_cn_short")
```

## **Andrea's comment**:
I checked the script that you mentioned (i.e. cycleways.R) and I have
one comment. The object osm_highway (line 8) includes all "OSM ways"
in the IoW and I think that the ratio nrow(osm_highway) /
nrow(osm_highways_cycling) does not represent a fair comparison (since
you are comparing ways and roads). If you want to estimate the ratio
of (cycleable streets) / (all streets in Iow), you can run:

osm_highway = osmextract::oe_get(place = "Isle of Wight", query =
"SELECT * FROM lines WHERE highway IS NOT NULL", quiet = TRUE)
osm_highways_cycling = osmextract::oe_get_network(place = "Isle of
Wight", mode = "cycling")

and you will notice that

nrow(osm_highway) / nrow(osm_highways_cycling) = 1

since, according to usual OSM tags, bicycle transportation is allowed
everywhere but for motorways, which are missing in the IoW. If you
want to create a more restrictive definition for a cycling network
(i.e. exclude primary or secondary roads), then you can modify the
code behind oe_get_network (see also the old PR:
https://github.com/ropensci/osmextract/pull/224).


```{r}
# applying Andrea's comment to West Yorkshare 

# osm_highway = osmextract::oe_get(place = region_name, query =
# "SELECT * FROM lines WHERE highway IS NOT NULL", quiet = TRUE, force_download = TRUE,
#                          force_vectortranslate = TRUE)
# osm_highways_cycling = osmextract::oe_get_network(place = region_name, mode = "cycling", force_download = TRUE,
#                          force_vectortranslate = TRUE)
nrow(osm_highway)
#> 180661
nrow(osm_highways_cycling)
#> 167853
c(nrow(osm_highway) == nrow(osm_highways_cycling), # FALSE
  nrow(osm_highway) / nrow(osm_highways_cycling) ) # 1.076305 ratio
# > 0.000000 1.076305

# checking if NULLs are NAs 
wy_cn_short %>% nrow()
#> 167853 
wy_cn_short %>% filter(!is.na(highway)) %>% nrow() # double checking if there are non NAs in the `highway` column
#> 167853 
wy_short %>% select(highway) %>% nrow() # includes ALL ways (inlcuding NA)
#> 233598
wy_short %>% filter(!is.na(highway)) %>% nrow() 
#> 180661

wy_short %>% filter(!is.na(highway)) %>% nrow() == nrow(osm_highway) # so NAs are NULLs as expected
# > TRUE

# Alternative way to figure out NAs
# wy_short %>% pull(highway) %>% is.na() %>% sum() # we have 52937 NAs
# wy_short %>% pull(highway) %>% table %>% sum() # 180661 NAs not included in the table
# ( (wy_short %>% select(highway) %>% nrow()) - (wy_short %>% pull(highway) %>% table %>% sum()) ) == wy_short %>% pull(highway) %>% is.na() %>% sum() # double checking

wy_short %>% filter(!is.na(highway)) %>% nrow() == wy_cn_short %>% nrow()  # we still have unequal number of rows; I think it might have something to do with highway cycling network definition
# > FALSE
(wy_short %>% filter(!is.na(highway)) %>% nrow()) - (wy_cn_short %>% nrow()) # we have 12808 more ways in wy_short compared to wy_cn_short
# > 12808

wy_short %>% pull(highway) %>% table()
wy_cn_short %>% pull(highway) %>% table()
# .
#      bridleway   bus_guideway         busway   construction       corridor       cycleway        footway  living_street       motorway  motorway_link             no           path     pedestrian 
#           1268             14              7            339             48           3250          36956            392            577            393              6           7286            801 
#        primary   primary_link       proposed        raceway    residential           road      secondary secondary_link        service       services          steps       tertiary  tertiary_link 
#           3415            321             78             17          44638             27           1470             43          53252              2           2685           5174            143 
#          track          trunk     trunk_link   unclassified 
#           7958           2888            338           6875 
# .
#      bridleway         busway       cycleway        footway  living_street       motorway  motorway_link           path     pedestrian        primary   primary_link    residential           road 
#           1260              6           3233          35543            368            571            389           6928            783           3415            321          44379             27 
#      secondary secondary_link        service       services       tertiary  tertiary_link          track          trunk     trunk_link   unclassified 
#           1470             43          46584              2           5156            142           7205           2878            338           6812 

# so I'll filter data based on the highway tags present in cycling network dataset `wy_cn_short`
highway_tags <- c("bridleway", "busway", "cycleway", "footway", "living_street", "motorway", "motorway_link", "path", "pedestrian", "primary", "primary_link", "residential", "road", "secondary", "secondary_link", "service", "services", "tertiary", "tertiary_link", "track", "trunk", "trunk_link",  "unclassified")
wy_checking <- wy_short %>% filter(highway %in% highway_tags)
wy_checking %>% nrow() 
# > 177474
wy_checking %>% nrow() == wy_cn_short %>% nrow() 
# > FALSE
( wy_checking %>% nrow() ) - ( wy_cn_short %>% nrow() ) # still significantly more lines are present in the wy_short dataset; cycling network must have had some more refined filtering; 
# > 9621



# wy_checking %>% select(highway) %>% plot(main = "wy_checking")
# wy_cn_short %>% select(highway) %>% plot(main = "wy_cn_checking")
```

It's interesting that Andrea's comment is applicable to IoW but not WY, I reckon it depends on the types of ways/roads present in the area and how well they are tagged. 

## keys and tags for cycling/wheelchair

tags of interest for cycling:
surface
bicycle
maxspeed:type
width
cycleway
cycleway:both
cycleway:right
cycleway:left
wheelchair
kerb
cycleway:both:lane
cycleway:left:lane
cycleway:right:lane
cycleway:left:width
cycleway:right:width
oneway:bicycle
cycleway:otherside                
cycleway:otherside:width
cycleway:both:width
maxwidth
cycleway:oneside
cycleway:oneside:width
cycleway:right:oneway
cycleway:width
cycleway:surface
width:lanes
bicycle_road
mobility_scooter
cyclestreet
cycleway:buffer
cycleway:left:separation:right
disabled
bicycle:lanes:conditional
bicycle:lanes:forward:conditional
cycleway:lane
cycleway:left:foot
cycleway:left:oneway                   
cycleway:left:segregated              
cycleway:left:seperation:right         
cycleway:proposed                      
cycleway:right:separation:left
handicap
"cycle"
"cycleway:est_width"

### data
```{r}
et <- c("surface", 
        "cycleway:surface",
        "bicycle", 
        "bicycle_road",
        "cyclestreet", # https://wiki.openstreetmap.org/wiki/Key:cyclestreet
        "cycleway",
        "maxspeed",
        "maxspeed:type",
        "width",
        "cycleway:width",
        "maxwidth",
        "cycleway:both", 
        "cycleway:right",
        "cycleway:left",
        "wheelchair",
        "disabled",
        "kerb"
        )

roads <- osmextract::oe_get("west yorkshire",
                            extra_tags = et,
                            force_vectortranslate = TRUE)
roads %>% glimpse()
wy %>% glimpse()
```

### key:maxspeed (needs recoding)

```{r}

? osmextract::oe_get_keys(wy_short)
all_tags <- osmextract::oe_get_keys(wy_short) 
all_tags


# maxspeed
all_tags %>% str_detect("maxspeed") 
speed = all_tags %>% str_detect("maxspeed")
wy_short[speed, ] %>% select(highway) %>% plot()

# speed limits
speed_data <- wy_short %>% mutate(speed = case_when(
  str_detect(other_tags,"10 mph") ~ 10,
  str_detect(other_tags,"20 mph") ~ 20,
  str_detect(other_tags,"30 mph") ~ 30,
  str_detect(other_tags,"40 mph") ~ 40,
  str_detect(other_tags,"50 mph") ~ 50,
  str_detect(other_tags,"60 mph") ~ 60,
  str_detect(other_tags,"70 mph") ~ 70,
  str_detect(other_tags,"80 mph") ~ 80,
  str_detect(other_tags,"90 mph") ~ 90,
  str_detect(other_tags,"100 mph") ~ 100)
                                           ) %>% 
  filter(!is.na(speed)) # getting rid of NAs

# how many rows
speed_data %>% nrow()
# > 19754
speed_data %>% pull(speed) %>% unique # double checking if all limits are present
speed_data %>% select(speed) %>% plot()

# leeds subset

# wy_pct = pct::get_pct_zones(region = "west-yorkshire")
# sf::sf_use_s2(FALSE)
# saveRDS(wy_pct,
#         "wy_pct.Rds")

# leeds_pct <- wy_pct %>% filter(lad_name == "Leeds")
# saveRDS(leeds,
#         "leeds_pct.Rds")


# plotting speed data on leeds
st_geometry(leeds) %>% plot(reset = FALSE, 
                            col = "gray",
                            main = "speed limits in Leeds (gray area)")
speed_data['speed'] %>% plot(add = TRUE, 
                             legend = TRUE
                             )

```



### key:surface
```{r}
# key:surface https://wiki.openstreetmap.org/wiki/Key:surface
roads %>% pull(surface) %>% table()%>% sort(decreasing = TRUE)  # for future: how to aggregate surface types?
roads %>% pull(surface) %>% table() %>% sum()
round((roads %>% pull(surface) %>% table() %>% sum() / (roads %>% nrow ) ) * 100, 2)
#> 12.33


roads1 <- roads %>% mutate(surface_coded = case_when(
  stringr::str_detect(surface,"asphalt") ~ "asphalt",
  stringr::str_detect(surface,"grass") ~ "grass",
  !stringr::str_detect(surface, c("asphalt","grass")) ~ "other" # values that aren't 'asphalt' or 'grass' are "other"
  ))

roads1 %>% pull(surface_coded) %>% table %>% sort(decreasing = TRUE)
#> asphalt   other   grass 
#> 14319   11835    2656  
roads1 %>% select(surface_coded) %>% plot()

```

### cycleway:surface
```{r}

roads1 %>% names()

round( ( roads1 %>% pull(cycleway_surface) %>% table() ) / ( roads1 %>% nrow() ) * 100, 3 )
#> 0.004

roads1 %>% select(cycleway_surface) %>% plot()

```


### key:bicycle
```{r}
# key:bicylce https://wiki.openstreetmap.org/wiki/Key:bicycle
roads1 %>% pull(bicycle) %>% table %>% sort(decreasing = TRUE)
roads1 %>% select(bicycle) %>% plot()

round((roads1 %>% pull(bicycle) %>% table() %>% sum() / (roads1 %>% nrow ) ) * 100, 2)
# > 2.95

# roads1 %>% select(highway) %>% plot(reset = FALSE,
#                                      col = 'black')
# roads1 %>% select(cycleway) %>% plot(add = TRUE,
#                                     # lwd = 2,
#                                      col = "red") # this is weird
```

### key:cycleway
```{r}
# key:cycleway https://wiki.openstreetmap.org/wiki/Key:cycleway
# Add the cycleway=* tag to a highway=* to map cycling infrastructure that is an inherent part of the road. This specifically applies to cycle lanes which are always a part of the road, and often also applies to separated cycle tracks if they are running parallel and next to the road. 
roads1 %>% pull(cycleway) %>% table %>% sort(decreasing = TRUE)
round((roads1 %>% pull(cycleway) %>% table() %>% sum() / (roads1 %>% nrow ) ) * 100, 2) # percentage of cycleways in relation to highways
#> 0.38

roads1 %>% select(cycleway) %>% plot() # plotting cycleways

st_geometry(leeds) %>% plot(reset = FALSE,
                            col = "gray")
roads1 %>% select(highway) %>% plot(add = TRUE,
                                     col = 'black')
roads1 %>% select(cycleway) %>% plot(add = TRUE,
                                     lwd = 2) # plotting cycleways in relation to highways 

# roads1 %>% select(bicycle) %>% plot(reset = FALSE,
#                                     col = "black")
# roads1 %>% select(cycleway) %>% plot(add = TRUE,
#                                      lwd = 2)
# roads1 %>% pull(bicycle) %>% table()
# 
# roads1 %>% select(highway) %>% plot(reset = FALSE,
#                                     col = "black")
# roads1 %>% select(bicycle) %>% plot(add = TRUE,
#                                     col = 'red')

# roads1  %>% filter(!is.na(bicycle)) %>% nrow()

# roads1 <- st_join(roads1, speed_data
#                 ) 




## tags
cl <- roads1 %>% select(cycleway_left)
cr <- roads1 %>% select(cycleway_right) 
cb <- roads1 %>% select(cycleway_both)


?plot

```


### key:width and cycleway:width
https://wiki.openstreetmap.org/wiki/Key:width
https://wiki.openstreetmap.org/wiki/Key:cycleway#Supplementary_details


```{r}
roads1 %>% names()
roads %>% pull(width) %>% table()
    #  -1     .25       0     0.1     0.3     0.4     0.5    0.5m     0.6     0.7    0.75     0.8     0.9       1 1 - 1.5     1.0     1.1     1.2     1.3 
    #   1       1      16       7      20      24      60       1      27       4       1      25       4     175       2       6       8      15       6 
    # 1.4     1.5    1.52     1.6     1.7    1.75     1.8     1.9      10    10.5      11      12    13.5      14      1m       2     2-3     2.0   2.117 
    #   1      40       3       9       3       3      11       2      10       1       2       1       1       1       2      89       1       3       1 
    # 2.2     2.3     2.4     2.5     2.6     2.7     2.8     2.9      2+      23       3     3.0     3.2     3.4     3.5   3.5 m     3.8    37.5      3m 
    #  10       4      11      77      11       3       1       3       1       4      97       2       7       2      12       1       2       1       1 
    #   4     4.5   4.5 m     4.8      46      4m       5       6     6.5      6m       7     7.4     7.5       8     8.5       9  narrow 
    #  41       2       1       1       3       1      17       4       2       1       3       1       1       9       3       2       9 
roads1 %>% pull(cycleway_width) %>% table()
 # 1.5 1.75 
 #   1    9 

roads1 %>% pull(width) %>% table() %>% sum()
#> 938

round((roads1 %>% pull(width) %>% table() %>% sum() / (roads1 %>% nrow ) ) * 100, 2)
#> 0.4
roads1 %>% select(width) %>% plot()

round((roads1 %>% pull(cycleway_width) %>% table() %>% sum() / (roads1 %>% nrow ) ) * 100, 3)


```

### key:cyclestreet

```{r}
roads1 %>% pull(cyclestreet) %>% table()
#> 2
round((roads1 %>% pull(cyclestreet) %>% table() %>% sum() / (roads1 %>% nrow ) ) * 100, 5)
#> 0.00086

```

### key:wheelchair
https://wiki.openstreetmap.org/wiki/Key:wheelchair
```{r}

roads1 %>% pull(wheelchair) %>% table()
       # bad designated    limited         no permissive    unknown        yes 
       #   2          3          3        113          2          1        194 
round((roads1 %>% pull(wheelchair) %>% table() %>% sum() / (roads1 %>% nrow ) ) * 100, 2)
#> 0.14
roads1 %>% select(wheelchair) %>% plot()

```

### key:disabled 
https://wiki.openstreetmap.org/wiki/Key:disabled
```{r}
roads1 %>% pull(disabled) %>% table()

#> 0.14
round((roads1 %>% pull(disabled) %>% table() %>% sum() / (roads1 %>% nrow ) ) * 100, 5)
#> 0.00086
roads1 %>% select(wheelchair) %>% plot()
```

### key:kerb
https://wiki.openstreetmap.org/wiki/Key:kerb

```{r}
roads1 %>% pull(kerb) %>% table()
     # fl   flush     low lowered  raised     yes 
     #  2      92       1      51      21       3 

#> 0.14
round((roads1 %>% pull(kerb) %>% table() %>% sum() / (roads1 %>% nrow ) ) * 100, 2)
#> 0.07

roads1 %>% select(kerb) %>% plot() # mostly indicated in Leeds?


st_geometry(leeds) %>% plot(reset = FALSE)
roads1 %>% select(kerb) %>% plot(add = TRUE,
                                 lwd = 4)
```


# Keys and tags for walking

```{r}
oe_get_keys(wy)

et_walking <- c("wheelchair",
"kerb",
"disabled",
"mobility_scooter",
"handicap",
"foot",
"lit", # https://wiki.openstreetmap.org/wiki/Key:lit
"access",
"sidewalk",
"footway",
"incline",
"smoothness",
"est_width",
"ramp",
"sidewalk_left",
"sidewalk_right",
"ramp_wheelchair",
"footway_left",
"footway_right",          
"footway_surface", 
"priority",
"sidewalk_both_surface", 
"path",                                   
"pedestrian",
"capacity_disabled",
"sidewalk_left_width",                    
"sidewalk_right_surface",
"width",
"tactile_paving")

oe_match_pattern("Yorkshire")
region_name <- "West Yorkshire"

wy_walking <- osmextract::oe_get(region_name,
                                 force_vectortranslate = TRUE,
                                 extra_tags = et_walking
                                 )
wy_walking %>% glimpse()
wy_walking %>% names()

# saveRDS(wy_walking,
#         "wy_walking.Rds")

# wy_walk_short <- wy_walking %>% select("foot",
#                 "lit",
#                 "access",
#                 "sidewalk",
#                 "footway",
#                 "incline",
#                 "smoothness",
#                 "est_width",
#                 "ramp",
#                 "sidewalk_left",
#                 "sidewalk_right",
#                 "ramp_wheelchair",
#                 "footway_left",
#                 "footway_right",          
#                 "footway_surface", 
#                 "priority",
#                 "sidewalk_both_surface", 
#                 "path",                                   
#                 "pedestrian",
#                 "capacity_disabled",
#                 "sidewalk_left_width",                    
#                 "sidewalk_right_surface")
# saveRDS(wy_walk_short,
#         "wy_walk_short.Rds")
wy_walk_short %>% glimpse
```




### tag:foot, footway, sidewalk

```{r}
# ============= FOOT
wy_walk_short %>% pull(foot) %>% table() 
  # customers    delivery  designated destination discouraged   emergency     limited          no  permissive      permit     private     unknown         yes 
         # 38           1        3652          34           3           1           1         374         516           1         123         151        7479

round((wy_walk_short %>% pull(foot) %>% table() %>% sum()) / (wy_walk_short %>% nrow()),2)
#> 0.05
wy_walk_short %>% select(foot) %>% plot()


# ============== FOOTWAY
wy_walk_short %>% pull(footway) %>% table()
 # access_aisle          alley       crossing           left           link             no           none             np       sidewalk traffic_island 
 #            77              1            505             11              1              9              3              1           1505              6 
 #           yes 
 #             5 
round((wy_walk_short %>% pull(footway) %>% table() %>% sum()) / (wy_walk_short %>% nrow()),2)
#> 0.01

# ============== SIDEWALK
wy_walk_short %>% pull(sidewalk) %>% table()
      # both   crossing       left left;right     mapped         no       none      right   separate        yes 
      #  820          1        431          1         23        452        163        200         79          6 
round((wy_walk_short %>% pull(sidewalk) %>% table() %>% sum()) / (wy_walk_short %>% nrow()),2)
#> 0.01


```


```{r}
osmextract::oe_get_keys(zone = "West Yorkshire")
```



## things to explore further:
- definition of cycling network in osm package
  - does it align with LTN 1/20 
  - key:cycleway in osm data
  - key:bicycle
- how 'access' column can be used to evaluate the quality of the cycling network?
- aggregate roads by speed limits?
  - overlap with cycling network
  - what does it say about the safety of cycling?
- how to aggregate surface types?
- not sure how cycleway:surface is tagged? has only 10 values. Any use?
- how to incorporate disability (e.g., wheelchair) related information?
  - small amount of data (maybe only an issue in WY?)
  - what are the street sign regulations for the disabled?
  - how do mappers evaluate (or even consider doing so?!) if its disabled-friendly
  - disability is not limited to wheelchair use but also blindness, etc.
  

















